---
title: "Run Example Decontamination"
author: "Sam Setta"
date: "`r Sys.Date()`"
output: html_document
knit: (function(inputFile, encoding) { 
  rmarkdown::render(inputFile, encoding = encoding, knit_root_dir = getwd(),
  output_file = file.path(dirname(inputFile), "run_example_partI_18Sv4.html")) 
  })
---

## Goal of Script

Set filtering method for ASV decontamination, steps we decided on:

**Part I Processing steps - THIS SCRIPT**

 * Step 1: Merge reads only (DONE)
 * Step 2: Tag jumping/index hopping
 * Step 3: Remove negative control contaminants.
 * Step 4: Remove the positive and negative control samples.
 * Step 5: Remove known contaminants (human, food products, pets, lab controls, etc.). Remove off-targets (i.e. taxonomic: bacteria/archaea in the MiFish; or length based). Leave in unknowns.

Produce Part I filtered output file

Following: https://github.com/zjgold/gruinard_decon

More details on how each works copied for the Readme:

**Estimation of Tag-jumping or Index Hopping**

With this procedure, we substract the proportion of reads that jumped into control samples from each environmental sample. First we determine which ASVs came from controls vs environmental samples. Next we create a vector of ASVs in positive controls. We then calculate what proportion of the reads found in the positive controls are found in environment samples. Next, we substract the composition of the positive controls from the environment samples. The idea behind this procedure is that we now know how many reads from each ASV appeared in the controls. These come from 2 processes: sequences we know should appear in the positive controls, and sequences that have hopped from the environment to the positive controls.

Load libraries

```{r setup, warning=FALSE, message=FALSE}
# install libraries
# BiocManager::install(c("tidyverse", "lubridate", "RColorBrewer", "ggsci", "dplyr", "phyloseq", "knitr", "readxl","ggplot2","NatParksPalettes","purrr", "ggvenn", "zetadiv"))

# devtools::install_github("donaldtmcknight/microDecon")

library("tidyverse")
library("proxy")
library("microDecon")
library("knitr")
library("vegan")
library("phyloseq")
library("DESeq2")
library("ape")
library("RColorBrewer")
library("ggsci")
library("readxl")
library("here")
library("openssl")
library("optparse")
library("configr")
library("Biostrings")
library("NatParksPalettes")
library("gt")
library("googlesheets4")
library("powerjoin")
here()

```

## Set-Up for decontamination

### Load decontam decontamination_utilities file
```{r load-R-utilities, warning=FALSE, message=FALSE}
# only run after confirming decontamination_utilities.R is in the script/decontamination folder
source(here("scripts","decontamination","decontamination_utilities.R"))
```

### Load config file for processing.

```{r load-config, warning=FALSE, message=FALSE}
# load config file of variables for script
option_list <- list(
  make_option(c("-c", "--config"), type="character", default=here("scripts","decontamination", "decontam_pre-merge_18Sv4.yaml"), help="Path to configuration file [default %default]", metavar="FILE"))

opt_parser <- OptionParser(option_list=option_list)
opt <- parse_args(opt_parser)
# check if yaml
is.yaml.file(opt$config)
eval.config.sections(opt$config)
config_file <- read.config(opt$config)
# what metabarcoding region are we working with?
config_file$region

```

### Load data

```{r load-data, echo=FALSE, warning=FALSE, message=FALSE}
# create output directories:
dir.create(here(config_file$output_path,config_file$region,"Output_csv"), recursive = TRUE)
dir.create(here(config_file$output_path,config_file$region,"Output_R"), recursive = TRUE)
dir.create(here(config_file$output_path,config_file$region,"Output_plots"), recursive = TRUE)
# load metadata file: (period in sample name rather than underscore)
metadata_file <- read_csv(
  here(config_file$input_meta_path)
) 
  
# load asv table: (underscore in sample name rather than period)
asv_table <- read.table(
  here(config_file$input_ASV_path),
  header = 1,
  sep = "\t",
  stringsAsFactors = F
)

# format asv table
if("seq_id" %in% asv_table) {
  asv_table_1 <- asv_table %>%
                      rename_with(~ str_remove(., "^MP_"))
  }else{
        asv_table_1 <- asv_table %>%
                      dplyr::rename("seq_id" = x) %>%
                      rename_with(~ str_remove(., "^MP_"))
}

# load fasta file
fasta_table <- readDNAStringSet(here(config_file$input_fasta_path),
  format =
    "fasta"
)

# format fasta file
fasta_df <- as.data.frame(fasta_table) %>%
  as_tibble(rownames = "ASV") %>%
  dplyr::rename("Sequence" = `x`) %>%
  mutate(seq_id = as.character(md5(Sequence)))

# load taxonomy file
taxonomy_table <- read.table(
  here(config_file$input_taxonomy_path),
  header = 1,
  sep = "\t",
  stringsAsFactors = F
) 

# format taxonomy table
if("Taxon" %in% taxonomy_table) {
  taxonomy_table_1 <- taxonomy_table %>%
                      dplyr::rename("ASV" = `Feature.ID`) %>%
                      left_join(fasta_df %>% select(c("ASV","seq_id"))) %>%
                      mutate(Taxon = str_remove(Taxon, ";$"))
  }else{
        taxonomy_table_1 <- taxonomy_table %>%
                      unite("Taxon", -ASV, sep=";", remove=TRUE) %>%
                      left_join(fasta_df %>% select(c("ASV","seq_id")))
}

```

### Create a dataframe of sequence lengths:
```{r seq-length-df, warning=FALSE}
# function to output sequence length from each file:
seq_len_df<-as.data.frame(fasta.seqlengths(here(config_file$input_fasta_path),
               nrec=-1L, skip=0L, 
               seek.first.rec=FALSE,
               seqtype="B", 
               use.names=TRUE)) %>%
    rownames_to_column(., var = "ASV") %>%
    rename_with(~ "Seq_length",2) %>%
    left_join(fasta_df %>% select(c("ASV","seq_id"))) %>% select(-"ASV")
# save file of sequence length
write_csv(seq_len_df, here(config_file$output_path,config_file$region,"Output_csv","Step0_seq_length.csv"))

```

### Format data

Metadata:

```{r format-metadata, echo=FALSE, warning=FALSE, message=FALSE}
# format metadata file
metadata <- metadata_file %>%
  mutate(region=case_when( # create region column to filter by for decontam.
        assay_name=="ssu16sv4v5_parada" ~ "Parada16S",
        assay_name=="ssu16sv4v5_parada_OSUmod" ~ "Parada16S",
        assay_name=="lsu16s_2434-2571_kelly" ~ "Kelly16S",
        assay_name=="ssu18sv4_stoeck" ~ "18Sv4",
        assay_name=="ssu18sv9_amaralzettler" ~ "18Sv9",
        assay_name=="ssu18sv8_machida_OSUmod" ~ "Machida18S",
        assay_name=="ssu18sv8_machida" ~ "Machida18S",
        assay_name=="COI_1835-2198_lerayfolmer" ~ "COI",
        assay_name=="ssu12sv5v6_mifish_u_sales" ~ "MiFish",
        assay_name=="ssu12sv5v6_mifish_u_sales_2xRSA" ~ "MiFish",
        assay_name=="ITS1_sterling" ~ "ITS1",
        assay_name=="dloopv1.5v4_baker" ~ "dLoop",
        assay_name=="lsu16s_2051-2438_mideca_komai" ~ "MiDeca"
  )) %>%
  mutate(expedition_id=case_when(
    expedition_id=="33RO20210613" | expedition_name=="WCOA21" ~ "WCOA21",
    TRUE ~ expedition_id
  )) %>%
  mutate(final_samp_name=samp_name, # keep original metadata sample name for final format
         samp_name=gsub("[.-]", "_", samp_name), # remove the cruise code after dash, since dada2 (revamp) output doesn't have this (ASV table)
         expedition_id=gsub("[.-]", "_", expedition_id), # replace dashes and periods with underscores for expedition ID
         cruise_yr=word(expedition_id, 1, sep = "_")) %>% # pattern to match asv sample names with cruise_yr id (first part of expedition id that needs to be added to sample_Name)
  mutate(perc_reads_retained=(output_read_count/input_read_count)*100) %>% # calculate percent reads retained (from raw reads to final ASV assignment)
  filter(region %in% config_file$region) %>% # only keep region of interest
  filter(seq_run_id %in% paste0("pmel_ome_", config_file$run)) %>% # only keep run of interest
  mutate(samp_name=str_replace(samp_name, fixed(paste0("_", expedition_id)), "")) %>%
  # remove any QIAvacTest from the samp_name for matching:
  mutate(samp_name=str_replace_all(samp_name, "_QIAvacTest\\d$", "")) %>%
  # remove any QIAvacTest from the samp_name for matching:
  mutate(samp_name=str_replace_all(samp_name, "(_QIAvacTest\\d$|_QIAvacTest)", ""),
         samp_name=str_replace_all(samp_name,"(_E$)", ""))

```

Merge metadata with asv table

```{r merge-data, echo=FALSE, warning=FALSE, message=FALSE}
# Generate Hash.key (essentially taxonomy table)
merged_df_matching <- asv_table_1 %>%
  pivot_longer(cols = -seq_id) %>% # transpose dataframe to edit sample name
  pivot_wider(names_from = seq_id, values_from = value) %>% dplyr::rename(samp_name="name") %>%
  mutate(asv_samp_name=samp_name, # keep original sample name for later
         samp_name=case_when(
              str_detect(samp_name, regex("POSITIVE", ignore_case = TRUE)) ~ samp_name, # keep POSITIVE sample ID
              str_detect(samp_name, regex("blank", ignore_case = TRUE)) ~ samp_name, # keep blank sample ID
              str_detect(samp_name, regex("pool", ignore_case = TRUE)) ~ samp_name, # keep blank sample ID
              TRUE ~ str_remove( # remove additional string from sampleID to remove expedition id to match metadata
           str_replace_all(samp_name, paste0(unique(metadata$cruise_yr), collapse = "|"), ""),"_$"))
           ) %>%
  mutate(samp_name=case_when(
    str_detect(samp_name, regex("pool", ignore_case = TRUE)) ~ samp_name,
    TRUE ~ str_remove(
    str_replace_all(samp_name, paste0(unique(metadata$expedition_name), collapse = "|"), ""),"_$"))
    )  %>%
  mutate(samp_name=case_when(
    str_detect(samp_name, regex("pool", ignore_case = TRUE)) ~ samp_name,
    TRUE ~ str_remove(
    str_replace_all(samp_name, paste0(unique(metadata$expedition_id), collapse = "|"), ""),"_$")) ) %>%
  # replace patterns that make it hard to match samp_names:
  mutate(samp_name=str_remove(str_replace_all(samp_name, 
  "(SKQ2021|MiFishMod|QiavacTest\\d$|QiavacTest|SKQ23_12S|Kotzebue23|Chaba22|12S)",""),"_$")) %>%
  mutate(samp_name=str_replace_all(samp_name, "__", "_")) %>%
  mutate(samp_name=str_replace_all(
    # rename positive sample
    samp_name, "POSITIVE",paste0(config_file$run,"_",config_file$region,"_POSITIVE"))) %>%
  mutate(samp_name=str_remove(str_replace_all(samp_name, "_\\d+(?=_|$)", "_"),"_$")) %>%
  mutate(samp_name=str_replace_all(samp_name, "__", "_")) %>%
  mutate(samp_name=str_replace_all(samp_name, "Mid_NC", "MID_NC")) %>%
  mutate(samp_name=str_replace_all(samp_name,"(_E$)", "")) %>%
  full_join(metadata %>% # join sample IDs with those from metadata
              select("samp_name","expedition_id","final_samp_name","region"), 
            by="samp_name") %>% 
  mutate(asv_samp_name_orig=asv_samp_name) %>%
  mutate(asv_samp_name = case_when( # OME WCOA21 samples only, do not have consistent naming, add the underscore WCOA21 to the E numbers
                str_detect(samp_name, regex("pool", ignore_case = TRUE)) ~ asv_samp_name,
                str_detect(final_samp_name, "WCOA21") ~ str_c(asv_samp_name, "_WCOA21"),
                TRUE ~ asv_samp_name)) 

# save original sample_ID and final_sample_name
if (!file.exists(here("data","processed","decontamination","sample_mismatch.csv"))) {
  # If it doesn't exist, create it
  write_csv(merged_df_matching %>% select(asv_samp_name=asv_samp_name_orig,final_samp_name,region), 
            here("data","processed","decontamination","sample_mismatch.csv"))
} else {
  # If it exists, read it, concat (bind), and save
  existing_df <- read_csv(here("data","processed","decontamination","sample_mismatch.csv"))
  combined_df <- bind_rows(existing_df, merged_df_matching %>%
                             select(asv_samp_name=asv_samp_name_orig,final_samp_name,region)) %>%
    distinct()
  write_csv(combined_df, here("data","processed","decontamination","sample_mismatch.csv"))
}

# re-format ASV table: merged_df_1
merged_df <- merged_df_matching %>% # only keep final sample name to convert back into asv table format
  select(-c("samp_name","expedition_id","region", "asv_samp_name","asv_samp_name_orig")) %>%
  dplyr::rename("samp_name"="final_samp_name") %>% # rename final sample name to keep for later processing
  relocate("samp_name") %>% # move samp_name back to beginning of dataframe
  pivot_longer(cols = -samp_name) %>% # transpose dataframe back into ASV table
  distinct() %>%
  pivot_wider(names_from = samp_name, values_from = value) %>% 
  dplyr::rename("ASV"="name") %>% # rename seq_id
  left_join(taxonomy_table_1) %>% # add taxonomy, seq_id(hash) to asv table
  left_join(fasta_df) %>% # add sequence to dataframe
  select(where(~ !all(is.na(.x)))) # filter out any empty columns (e.g. if there is no positive control)

matching_lookup <- merged_df_matching %>% 
  dplyr::select(c("asv_samp_name","final_samp_name")) %>%
  filter(!(is.na(asv_samp_name) & str_detect(final_samp_name, regex("POSITIVE", ignore_case = TRUE))))


# Keep only ASV and taxonomy columns
hash_key <- merged_df %>%
  dplyr::select(seq_id, Taxon)

# save metabarcoding region id in merged dataframe:
merged_df$Miseq_region <- config_file$region
```

Add separated taxonomic levels if not already in dataframe

```{r taxa-format, warning=FALSE, message=FALSE}
## Add Taxonomy if missing:
if("Species" %in% names(hash_key)) {
  message("Taxonomy already separated — skipping formatting step")
} else {
        if (config_file$taxonomy_format=="nine_level") {
        hash_key <- separate_wider_delim(hash_key, Taxon, 
                             names= c("Domain", "Supergroup", "Division", "Subdivision", "Class", 
                                      "Order", "Family", "Genus", "Species"), 
                             delim=";", too_few="align_start", too_many="drop", cols_remove = FALSE)
        } else {
          hash_key <- separate_wider_delim(hash_key, Taxon, 
                             names= c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
                             delim=";", too_few="align_start", too_many="drop", cols_remove = FALSE) }
}
```

### Merge tables (if using more than one and check format)

```{r check-merge-format, warning=FALSE, message=FALSE}
# Merge All Tables (if using Anacapa taxonomy, but here, just based off revamp)
merged_table <- merged_df
```

### Convert to long format:

```{r convert-to-long, echo=FALSE, warning=FALSE, message=FALSE}
### Format for Long Data
merged_table$seq_id <- factor(merged_table$seq_id)
merged_table$Miseq_region <- factor(merged_table$Miseq_region)

# save column names from merged_table:
columns <- colnames(merged_table)
# create list of column names to remove
remove <- c(
  "ASV",
  "Taxon",
  "Miseq_region",
  "seq_id",
  "Sequence",
  "Miseq_region"
)

# columns to keep via inverse filtering (all those not found in the 'remove' list)
gathercols <- columns[!columns %in% remove]

# create long dataframe by using only the sample IDs (those not found in 'remove' or only those found in 'gathercols')
merged_table_long <- merged_table %>%
  # select(-remove) %>%
  # column_to_rownames("ASV") %>%
  pivot_longer(
    .,
    cols = all_of(gathercols),
    names_to = "sample",
    values_to = "reads"
  )

# save reads as numeric
merged_table_long$reads <- as.numeric(merged_table_long$reads)

```

## Step 1: Merge reads only (DONE)

Need to change this if using anacapa results.

## Step 2: Tag jumping/index hopping 

### 2.1 Nest the dataset by origin of ASVs:

#### Label ASVs identified in positive control, negative control, or Samples

Identify Positives, Negatives, and Samples

```{r id-controls, warning=FALSE, message=FALSE}
# save percent reads retained as vector:
reads_retained_con<-metadata %>%
  dplyr::select("final_samp_name","perc_reads_retained","neg_cont_type","samp_category") 

### New column that labels each ASV as from Positive (control) or Sample
merged_table_long_v2 <-
  merged_table_long %>%
  left_join(reads_retained_con, by = join_by("sample"=="final_samp_name")) %>% # add percent reads retained and control types
  filter(reads != 0) # removing empty sequences here bc long format, so just adds more computational time here to echo.

```

Create positive control filters:

- % of raw reads retained greater than 30
- Total reads > 30000

- Flag 1:  “Pass” perc_reads_retained > 30 & TotalReadsperSample > 30000
- Flag 2: “TooFewReads” perc_reads_retained > 30 & TotalReadsperSample <= 30000
- Flag 2: “LowReadsRetained” perc_reads_retained <= 30 & TotalReadsperSample > 30000
- Flag 3: “Fail” perc_reads_retained <= 30 & TotalReadsperSample <= 30000

```{r pos-control-filters, echo=FALSE, warning=FALSE, message=FALSE}
# Check if there is a positive control in samples
if ("positive control" %in% merged_table_long_v2$samp_category && config_file$region!="18Sv4") {
  message("Positive control found — running code")
    # create a dataframe to use for positive control tag jumping decision later:
    pass_pos_filter_df <- merged_table_long_v2 %>%
      filter(samp_category=="positive control") %>%
      group_by(sample, samp_category, perc_reads_retained) %>%
      summarise(TotalReadsperSample = sum(reads)) %>%
      mutate(Pos_pass_filters=case_when(
        perc_reads_retained > 30 & TotalReadsperSample > 30000 ~ "pass",
        perc_reads_retained > 30 & TotalReadsperSample <= 30000 ~ "TooFewReads",
        perc_reads_retained <= 30 & TotalReadsperSample > 30000 ~ "LowReadsRetained",
        perc_reads_retained <= 30 & TotalReadsperSample <= 30000 ~ "fail")) %>%
      mutate(pos_flag=case_when(
        Pos_pass_filters=="pass" ~ "1",
        Pos_pass_filters=="TooFewReads" ~ "2",
        Pos_pass_filters=="LowReadsRetained" ~ "2",
        Pos_pass_filters=="fail" ~ "3",
      )) %>%
      dplyr::select(c(sample,Pos_pass_filters,pos_flag))
    
    # merge df and pos_filter_df:
    merged_table_long_v2 <- merged_table_long_v2 %>%
      left_join(pass_pos_filter_df)

  } else { # If there isn't a positive control output message
    message("Positive control not found — skipping analysis")

     merged_table_long_v2 <- merged_table_long_v2 %>%
      mutate(Pos_pass_filters=rep(NA,length(Miseq_region)),
              pos_flag=rep(NA,length(Miseq_region)))

  }

```

Load in positive control contaminants if 18Sv4:

```{r ome-pos-control-contaminants, echo=FALSE, warning=FALSE, message=FALSE}
# save parent directory to load in 18Sv4 positives
path_to_18Sv4_pos <- dirname(getwd())
# Check if there is a positive control in samples
if (config_file$region=="18Sv4" && 
  !is.null(config_file$OME_18Sv4_pos)) {
  message("Positive control for 18Sv4 loaded seperately")
    OME_18Sv4_pos <- read_csv(here(path_to_18Sv4_pos,config_file$OME_18Sv4_pos)) %>%
      filter(run==config_file$run)

    } else { # If there isn't a positive control output message
      message("Positive control not found — skipping analysis")
  }

```


Load known lab contaminants:

```{r load-known-contaminants, warning=FALSE, message=FALSE}
### Filter by Taxonomic Paths
# This list can be expanded upon additional work to identify known contaminants within OME PMEL samples
if (is.null(config_file$step5_species_to_remove_path)) {
  step5_species_to_remove <- tibble(taxa_level=c("Species","Species","Order","Species",
                                                    "Genus","Genus","Genus","Genus"),
                                       taxa_id=c(
                                         "Harpagornis moorei","Pan paniscus","Primates","Sus scrofa",
                                         "Bos","Homo","Gallus","Canis"))
} else {
  step5_species_to_remove <- read_csv(here(config_file$step5_species_to_remove_path))
}

```

Load negative control associated samples:

```{r load-neg-associated-samples, echo=FALSE, warning=FALSE, message=FALSE}
### Filter by Taxonomic Paths
if (is.null(config_file$step3_assoc_nc)) {
  step3_associated_nc <- NULL
} else {
  # Read in each sheet
  step3_associated_file <- read_csv(here(config_file$step3_assoc_nc)) %>%
                            filter(str_detect(associated_run, config_file$run)) # only keep associated nc's from run of interest

  # negative control filtered df:
  step3_associated_nc <- step3_associated_file %>% 
    filter(sequenced==TRUE) %>% # filter out only negative controls we would find in sequenced data
    filter(str_detect(associated_run, config_file$run)) %>% # only keep associated nc's from run of interest
    filter(str_detect(region_amplified, config_file$region))

  # df from saving negative control samples run in multiple sequencing runs:
  associated_multiple_seq_runs<-step3_associated_file %>%
    filter(str_detect(associated_run, ",") & sequenced==TRUE)
}

```

Load negative controls from other runs

```{r neg-controls-across-runs, echo=FALSE, warning=FALSE, message=FALSE}
# Only run negative control analysis if there's a negative control in multiple runs:
if (file.exists(config_file$step3_nc_across_runs)) {
  # load file of reads from other runs:
  nc_across_runs_input <- read_delim(here(config_file$step3_nc_across_runs)) %>%
    distinct() %>% # only keep unique rows because file is concatenated
    mutate(neg_flag=as.character(neg_flag))
  # add negative controls to step3_associated_nc if not already in file 
  nc_to_merge_with_associated <- nc_across_runs_input %>%
    dplyr::rename(region_amplified=Miseq_region) %>%
    # keep only associated run that matches run of interest:
    filter(str_detect(region_amplified, config_file$region)) %>%
    distinct()# remove any repeats because seq_id's were removed

  # save sample_id, reads, metadata to join later in processing:
  nc_add_to_merged_table_long_v2<-nc_across_runs_input %>%
    # keep only associated run that matches run of interest:
    filter(str_detect(Miseq_region, config_file$region)) %>%
    select("sample","Taxon","seq_id":"neg_flag") %>% 
    mutate(neg_flag=as.character(neg_flag)) %>% # make sure flag matches dataframe to merge with
    distinct() # remove any repeats after columns filtered out
  } else { 
    message("Skip - no additional negative controls from other runs to add")
}

```

#### Format associated negative control samples

```{r format-assoc-neg-controls, echo=FALSE, warning=FALSE, message=FALSE}
# Only run negative control analysis if there's a negative control in samples:
if ("negative control" %in% merged_table_long_v2$samp_category | !is.null(step3_associated_nc)) {
  message("Negative control found — running code")
  if ("pool" %in% step3_associated_nc$pool_single) {
    message("pooled samples - adding associated samples from source")
    # Separate pools and individuals
    pools <- step3_associated_nc %>% filter(pool_single == "pool") %>%
      # Split the associated_sample into long format
      separate_rows(pool_samp_source, sep = "[,]+") %>%
      # only keep columns needed for matching
      select("samp_name","pool_samp_source")
  
    # datasheet to match with:
    single_samples  <- step3_associated_nc %>% filter(pool_single == "single")   
      
    # Process pools separately
    pools_updated <- pools  %>%
      # Join with the individuals to get the values
      left_join(single_samples %>% select(-"pool_samp_source"), by = c("pool_samp_source" = "samp_name")) %>%
      # Group back by the original pool name and collapse
      group_by(across(-c("associated_field_sample","associated_extract_sample","pool_samp_source","date_of_extraction","date_of_collection_utc","note","region_amplified"))) %>%
      summarize(
        associated_extract_sample = paste(associated_extract_sample, collapse = ","),
        associated_field_sample = paste(associated_field_sample, collapse = ","),
        .groups = "drop"
      ) %>% 
      mutate(pool_single="pool", region_amplified=config_file$region) %>%
      filter(!is.na(associated_run))     
    
    # combine the two dataframes
    step3_associated_nc_1<-bind_rows(single_samples,pools_updated)
    
  } else {
    message("no pooled samples - continue to formatting")
    step3_associated_nc_1<-step3_associated_nc
  }
  # join pooled and single samples:
  step3_associated_nc_2 <- step3_associated_nc_1 %>%
    mutate(samp_name=gsub("[.-]", "_", samp_name)) %>%
    left_join(matching_lookup, by = c("samp_name" = "asv_samp_name")) %>%
    # START CONDITIONAL STEP
    {
      # Check if the object exists in the environment
      if (exists("nc_to_merge_with_associated")) {
        # Identify rows that didn't match and try joining with df3
        # We use left_join again, but we only want to fill in the NAs
        left_join(., 
                  nc_to_merge_with_associated %>% 
                    select(asv_samp_name, final_samp_name_other_runs = sample) %>% 
                    distinct(), 
                  by = c("samp_name" = "asv_samp_name")) %>%
          
          mutate(final_samp_name = coalesce(final_samp_name, final_samp_name_other_runs)) %>%
          select(-final_samp_name_other_runs) # Cleanup temporary columns
      } else {
        # If it doesn't exist, just pass the data through to the next step
        message("nc_to_merge_with_associated not found - skipping join.")
        . # keeps original dataframe
      }
    } %>% # END CONDITIONAL STEP
    dplyr::rename(sample=final_samp_name) %>%
    select(-c("samp_name")) %>% relocate("sample") %>%
    # separate associated field sample id's and rename with final sample names before collapsing 
    # into one column again for associated sample filtering.
    separate_rows(associated_field_sample, sep = "[,]+") %>% 
    mutate(associated_field_sample=gsub("[.-]", "_", associated_field_sample)) %>%
    left_join(matching_lookup, by=c("associated_field_sample"="asv_samp_name")) %>%
    group_by(across(-c("associated_field_sample","final_samp_name"))) %>%
    summarize(associated_field_sample = if(all(is.na(final_samp_name))) {
        NA_character_ # if only NAs, keep associated_field_sample column as NA
      } else { paste(final_samp_name[!is.na(final_samp_name)], collapse = ",") }, .groups = "drop") %>% # remove any NA values from final_samp_name
    # separate associated extracted sample id's and rename with final sample names before collapsing 
    # into one column again for associated sample filtering.
    separate_rows(associated_extract_sample, sep = "[,]+") %>% 
    mutate(associated_extract_sample=gsub("[.-]", "_", associated_extract_sample)) %>%
    left_join(matching_lookup, by=c("associated_extract_sample"="asv_samp_name")) %>%
    # filter(!is.na(final_samp_name)) %>% # CREATES AN ISSUE IF NOT PRESENT IN DATA!!!
    group_by(across(-c("associated_extract_sample","final_samp_name"))) %>%
    summarize(associated_extract_sample = if(all(is.na(final_samp_name))) {
        NA_character_ # if only NAs, keep associated_extract_sample column as NA
      } else { paste(final_samp_name[!is.na(final_samp_name)], collapse = ",") }, .groups = "drop")
    
    if(exists("nc_to_merge_with_associated")) {
      step3_associated_nc_final <- step3_associated_nc_2 %>%
        dplyr::select(-region_amplified) %>% # keep only the region of interest for this script
        full_join(nc_to_merge_with_associated, by=c("sample")) %>% 
        unite("all_associated_samples", associated_field_sample, associated_extract_sample, sep = ",",
            na.rm = TRUE, remove = FALSE) %>%
      mutate(all_associated_samples = na_if(all_associated_samples, "")) # remove empty strings created by unite if two NA columns
      
    } else {
    step3_associated_nc_final <- step3_associated_nc_2  %>%
      unite("all_associated_samples", associated_field_sample, associated_extract_sample, sep = ",",
            na.rm = TRUE, remove = FALSE) %>%
      mutate(all_associated_samples = na_if(all_associated_samples, "")) %>% # remove empty strings created by unite if two NA columns
      filter(!is.na(sample))
    }

  } else {
  message("Negative control not found or NC has no associated sample list")
  }

```

Add negative controls from other runs to merged table:

```{r add-neg-controls-other-runs, echo=FALSE, warning=FALSE, message=FALSE}
# Only run negative control analysis if there's a negative control in multiple runs:
if (file.exists(config_file$step3_nc_across_runs)) {
  message("Negative controls from other runs to be added")
  # join will full dataframe:
  # join will full dataframe:
  merged_table_long_v2 <-merged_table_long_v2 %>% 
    full_join(nc_add_to_merged_table_long_v2) %>%
    full_join(step3_associated_nc_final)
 
  } else {
    message("No additional negative controls from other runs to add")
    if ("negative control" %in% merged_table_long_v2$samp_category | !is.null(step3_associated_nc)) {
      message("Negative control found and associated samples added to full dataframe")
      # join will full dataframe:
      merged_table_long_v2 <- merged_table_long_v2 %>% 
        full_join(step3_associated_nc_final)
    } else {
      message("No associated negative control - skipping")
    }
  }

```

Separate taxonomy from dataframe:

```{r sep-taxa-if-needed, echo=FALSE, warning=FALSE, message=FALSE}
## Add Taxonomy if missing:
if("Species" %in% names(merged_table_long_v2)) {
  message("Taxonomy already separated — skipping formatting step")
} else {
        if (config_file$taxonomy_format=="nine_level") {
        merged_table_long_v2 <- separate_wider_delim(merged_table_long_v2, Taxon, 
                             names= c("Domain", "Supergroup", "Division", "Subdivision", "Class", 
                                      "Order", "Family", "Genus", "Species"), 
                             delim=";", too_few="align_start", too_many="drop", cols_remove = FALSE)
        } else {
          merged_table_long_v2 <- separate_wider_delim(merged_table_long_v2, Taxon, 
                             names= c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
                             delim=";", too_few="align_start", too_many="drop", cols_remove = FALSE) }
}

```

Create negative control filters:

- Flag 1: Off target seqs (12S seq length > 210) & TotalReadsperSample < 1000
- Flag 2: Sequences from known lab contaminants or read counts 1000 < TotalReadsperSample < 5000
- Flag 3: TotalReadsperSample > 5000 or anything other than lab contaminants

```{r create-neg-filters, echo=FALSE, warning=FALSE, message=FALSE}
# Check if there is a negative control in samples
if ("negative control" %in% merged_table_long_v2$samp_category) {
  message("Negative control found — running code")
  # create a check to see if there are any missing neg_flag's for negative controls 
  # if bringing in associated nc's from other run's there won't be.
  check <- "neg_flag" %in% names(merged_table_long_v2) &&
    merged_table_long_v2 %>% 
           filter(samp_category == "negative control") %>% 
           summarize(any_na = any(is.na(neg_flag))) %>% 
           pull(any_na)
  # run main part of code to create flags for negative controls:
  if(check | !("neg_flag" %in% names(merged_table_long_v2))) {
        if("neg_flag" %in% names(merged_table_long_v2)) {
        # create a dataframe
        neg_sample_flag_df <- merged_table_long_v2 %>%
          filter(samp_category=="negative control" & is.na(neg_flag)) 
            } else {
        neg_sample_flag_df <- merged_table_long_v2 %>%
          filter(samp_category=="negative control") 
          }
    # run for samples that need neg_flag:
    neg_sample_flag_df <- neg_sample_flag_df %>% # filter out negative controls
      left_join(seq_len_df, by="seq_id") %>% # add sequence length for negative control flags
      mutate(off_target_length=case_when(
        config_file$seq_length_filt=="Yes" & Seq_length > config_file$seq_length_max ~ "Off Target", 
        TRUE ~ "Not off target"
      )) %>% 
      mutate(known_contaminant=rep(FALSE, length(Miseq_region))) # Create a logical vector of FALSEs
     # Loop through the rules in your contaminant dataframe
     for (i in seq_len(nrow(step5_species_to_remove))) {
       level <- step5_species_to_remove$taxa_level[i]
       id <- step5_species_to_remove$taxa_id[i]
       # Update the logical vector: if it was already TRUE, keep it TRUE;
       # If logical vector was FALSE, keep FALSE, if NA, keep FALSE by using %in%
       # If TRUE already and next value is NA (bc taxa_level not id'd), keep TRUE
       neg_sample_flag_df <- neg_sample_flag_df %>%
         mutate(known_contaminant=known_contaminant | (neg_sample_flag_df[[level]] %in% id))
       }
      # Summarise and add flags:
      neg_sample_flag_df <- neg_sample_flag_df %>%
        group_by(sample, samp_category) %>%
        summarise(
                  Unknown_contaminants = any(known_contaminant==FALSE),
                  Not_off_target = any(off_target_length=="Not off target"),
                  TotalReadsperSample = sum(reads)) %>%  # summarise total reads
        mutate(neg_total_reads=TotalReadsperSample) %>%
        mutate(neg_flag=case_when(
          # Flag 1: Off target seqs (12S seq length > 210) & TotalReadsperSample < 1000 (whatever is in step2_low_read_depth)
          Not_off_target=="FALSE" & Unknown_contaminants=="FALSE" | 
            TotalReadsperSample < config_file$step2_low_read_depth & Unknown_contaminants=="FALSE"  ~ "1",
          # Flag 2: Sequences from known lab contaminants or read counts step2_low_read_depth < TotalReadsperSample < step2_high_read_depth
          Not_off_target=="TRUE" & Unknown_contaminants=="FALSE" |
          Not_off_target=="FALSE" & Unknown_contaminants=="FALSE" & 
            TotalReadsperSample >= config_file$step2_low_read_depth & 
            TotalReadsperSample < config_file$step2_high_read_depth ~ "2",
          # Flag 3: TotalReadsperSample > step2_high_read_depth or anything other than lab contaminants
          Unknown_contaminants=="TRUE" | TotalReadsperSample >= config_file$step2_high_read_depth ~ "3"))
    # merge df and pos_filter_df:
    merged_table_long_v2 <- merged_table_long_v2 %>%
      left_join(neg_sample_flag_df %>% select(-"TotalReadsperSample"))
  } else {
    message("Skip making negative flags - all from another run")
  }
} else { # If there isn't a negative control output message
    message("Negative control not found — skipping analysis")
}

```

Save negative controls associated with other samples

```{r save-neg-controls-other-runs, echo=FALSE, warning=FALSE, message=FALSE}
# Only run negative control analysis if there's a negative control in samples:
if (exists("associated_multiple_seq_runs")) {
  message("Negative control found — running code")
  associated_multiple_seq_runs_1 <- associated_multiple_seq_runs %>% 
    # correct sample naming:
    left_join(matching_lookup %>% 
                mutate(samp_name=str_replace_all(asv_samp_name,"_",".")), by="samp_name") %>%
    # keep only sample as the id column
    mutate(sample=final_samp_name) %>%
    dplyr::select(-c("samp_name","final_samp_name")) %>%
    filter(!is.na(sample))
  
  # keep only samples across multiple sequencing runs:
  associated_multiple_seq_runs_2 <- merged_table_long_v2 %>% 
    dplyr::select(-c("ASV":"Species","associated_run":"associated_extract_sample")) %>%
    left_join(associated_multiple_seq_runs_1 %>% select(sample, asv_samp_name)) %>%
    filter(sample %in% associated_multiple_seq_runs_1$sample) %>%
    relocate(asv_samp_name, .after = samp_category)
  # if negative control sequenced with this run, save in fill under decontam file:
  if(any(!is.na(associated_multiple_seq_runs_2$seq_id))) {
    write_delim(associated_multiple_seq_runs_2,
                file = here(config_file$output_path,config_file$region,"Output_csv","nc_associated_multiple_runs.txt"))
  } else {
    message("no negative control samples across multiple runs")
  }
} else {
  message("no associated negative controls")
}

```

#### Visualize Read Counts Across Samples for Barcode_1

```{r read-count-fig, warning=FALSE, message=FALSE}
plot_1 <- merged_table_long_v2 %>% 
  filter(if(config_file$region == "18Sv4") samp_category != "positive control" else TRUE) %>%
  group_by(sample) %>%
  mutate(TotalReadsperSample = sum(reads)) %>%
  arrange(desc(TotalReadsperSample)) %>%
  ggplot(., aes(x = sample, y = TotalReadsperSample, color = samp_category)) +
  geom_point() +
  ggtitle("Read Count Across Samples") +
  theme(axis.text.x = element_text(angle = 90))

plot_1

ggsave(
  plot = plot_1,
  here(config_file$output_path,config_file$region,"Output_plots","step2_Sample_Read_Depth_Barcode_1.png"),
  device = "png",
  width = 12,
  height = 8,
  units = "in"
)

```

#### Visualize Read Counts Across sample types

```{r read-count-table, echo=FALSE, warning=FALSE, message=FALSE}
# count the number of reads per sample type (positive control, negative control, sample)
# filter differently based on whether negative control samples exist in data:
if ("negative control" %in% merged_table_long_v2$samp_category) {
  message("Negative control found — keeping single samples only")
  table_1 <- merged_table_long_v2 %>%
  group_by(samp_category, sample) %>%
  summarise(TotalReadsperSample = sum(reads)) %>%
  ungroup() %>% group_by(samp_category) %>%
  summarise(Mean=round(mean(TotalReadsperSample),digits = 0),
            SD=round(sd(TotalReadsperSample),digits=0),
            Max=max(TotalReadsperSample),
            Min=min(TotalReadsperSample),
            Median=round(median(TotalReadsperSample),digits=0),
            n=n())
} else {
  table_1 <- merged_table_long_v2 %>%
  group_by(samp_category, sample) %>%
  summarise(TotalReadsperSample = sum(reads)) %>%
  ungroup() %>% group_by(samp_category) %>%
  summarise(Mean=round(mean(TotalReadsperSample),digits = 0),
            SD=round(sd(TotalReadsperSample),digits=0),
            Max=max(TotalReadsperSample),
            Min=min(TotalReadsperSample),
            Median=round(median(TotalReadsperSample),digits=0),
            n=n())
}

# nicer version of table to save
step2_summary_table_df <- table_1 %>% 
  filter(if(config_file$region == "18Sv4") samp_category != "positive control" else TRUE) %>%
  gt() %>%
  cols_label(
    samp_category = "Sample Type") %>%
  tab_style(
    style = list(
      cell_text(weight = "bold", size = "large", align = "center")
    ), locations = cells_column_labels()) %>%
   tab_style(
    style = list(
      cell_text(weight = "bold", size = "large", align = "center")
    ), locations = cells_column_labels()) %>%
  tab_style(
    style = cell_borders(
      sides = "all", 
      color = "black", 
      weight = px(1)
    ),
    locations = list(
      cells_body(),
      cells_column_labels(),
      cells_title()
    )
  ) %>%
  tab_options(
    table.border.top.color = "black",
    table.border.top.width = px(1),
    table.border.bottom.color = "black",
    table.border.bottom.width = px(1),
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black"
  )

# check table output:
step2_summary_table_df

# save table before filtering starts:
gtsave(step2_summary_table_df, here(config_file$output_path,config_file$region,"Output_plots","step0_sample_table.html"))
# save summary table as csv:
write_csv(table_1, file = here(config_file$output_path,config_file$region,"Output_csv","step2_summary_table_before_filt.csv"))

```

#### ASVs in positive controls

Beginning of code below will skip analysis if there are no positive controls present in the data.

```{r id-pos-control-sp, echo=FALSE, warning=FALSE, message=FALSE}
# Check if there is a positive control in samples
if ("positive control" %in% merged_table_long_v2$samp_category && config_file$region!="18Sv4") {
  message("Positive control found — running code")
  ## Summarise total number of reads per seq_id
  all_seqs_in_positives <- merged_table_long_v2 %>%
    filter(samp_category == "positive control") %>%
    dplyr::group_by(seq_id) %>%
    dplyr::summarise(tot = sum(reads)) %>%
    arrange(desc(tot)) 
  
  ## Pull in associated taxonomy for positive ASVs
  pos_contam_species <- all_seqs_in_positives %>%
    left_join(hash_key, by="seq_id") %>%
    as_tibble() 
  
  ## Save ASVs, taxonomy and read counts for positive control samples
  write_csv(pos_contam_species, 
            file = here(config_file$output_path,config_file$region,"Output_csv","step2_pos_contam_species.csv")
            )
} else { # If there isn't a positive control output message
  message("Positive control not found — skipping analysis")
}
```

#### ASVs in negative controls.

Beginning of code below will skip analysis if there are no negative controls present in the data.

```{r blank-contam-sp, echo=FALSE, warning=FALSE, message=FALSE}
# Check if there is a positive control in samples
if ("negative control" %in% merged_table_long_v2$samp_category) {
  message("Negative control found — running code")

  ## Summarise total number of reads per ASV
  all_seqs_in_blanks <- merged_table_long_v2 %>%
    filter(samp_category == "negative control") %>%
    dplyr::group_by(neg_cont_type, seq_id) %>%
    dplyr::summarise(tot = sum(reads)) %>%
    arrange(desc(tot))
  
  ## Pull in associated taxonomy for positive ASVs
  blank_contam_species <- all_seqs_in_blanks %>%
    left_join(hash_key, by="seq_id")
  
  ## Save ASVs, taxonomy and read counts for negative control samples
  write_csv(blank_contam_species, here(config_file$output_path,config_file$region,"Output_csv","step2_blank_contam_species.csv"))
} else { # If there isn't a positive control output message
  message("Negative control not found — skipping analysis")
}

```

#### Generate nested ASV table

```{r generate-nested-asv, echo=FALSE, warning=FALSE, message=FALSE}
## Generated ASV_nested
ASV_nested<- merged_table_long_v2 %>% 
    filter(Miseq_region==config_file$region) %>%
    dplyr::group_by(Miseq_region, samp_category) %>%
    nest() %>%
    pivot_wider(names_from = samp_category, values_from = data)
  
### Summary_file_1 
ASV_summary <- ASV_nested %>%
    ungroup() %>%
    dplyr::mutate(., Miseq_region, Summary = purrr::map(sample, ~ how.many(ASVtable = ., round = 0)), .keep = "none")
  ASV_summary$Summary

```

### 2.2: Model the composition of the positive controls of each run

```{r pos-control-comp-model, echo=FALSE, warning=FALSE, message=FALSE}
# Only run tag jumping in and out of positive controls if positive control in samples:
if (("positive control" %in% merged_table_long_v2$samp_category) &&
     ("pass" %in% merged_table_long_v2$Pos_pass_filters) &&
      config_file$region!="18Sv4") {
  message("Positive control found & passed filters — running code")
  #### Jumping vector out of Positive
ASV_nested <- ASV_nested %>%
    mutate(contamout_tibble = purrr::map(sample, function(.x) { # started new nested dataframe in R
      df <- .x %>%
        ungroup() %>%
        group_by(sample) %>%
        mutate(TotalReadsperSample = sum(reads))  %>% # calculate total reads per sample
        mutate(proportion = reads / TotalReadsperSample) %>% # calculate proportion of reads per sample
    # remove known positive control species and other higher level taxonomic:
    filter(., Taxon == config_file$step2_positive_species_taxonomic_path | # with species
                 !!sym(config_file$step2_positive_lowest_taxonomic_lvl) == config_file$step2_positive_lowest_taxonomic_lvl_id) %>% # highest level taxonomy of positive 
        # dplyr::select(-c(Domain:Species)) %>%
        ungroup() %>%
        group_by(seq_id)
      # if positive contamination species are not found in samples (no rows, add NA values)
      if (nrow(df) == 0) {
      tibble(seq_id = factor(NA, levels = levels(.x$seq_id)), vector_contamination = NA_real_)
      # if positive contamination species are found in samples calculate maximum proportions for vector contamination
    } else {
      df %>%
        summarise(vector_contamination = max(proportion, na.rm = FALSE), .groups = "drop")
    }
    }))
      ### Jumping vector into Positive
       ASV_nested <- ASV_nested %>%
        mutate(contam_tibble = purrr::map(`positive control`, function(.x) { # create new nested dataframe using positive nested dataframe created above
          .x %>%
            ungroup() %>%
            group_by(sample) %>%
            mutate(TotalReadsperSample = sum(reads)) %>%  # calculate total reads per sample
            mutate(proportion = reads / TotalReadsperSample) %>% # calculate proportion of reads per sample
            group_by(seq_id) %>% # this can be changed to Taxon
            dplyr::summarise(vector_contamination = max(proportion)) # calculate maximum proportions for vector contamination
        }))

    ### Vector Contamination in Barcode_1 visualization
    vc_plot_1<- rbind(ASV_nested$contam_tibble[[1]], ASV_nested$contamout_tibble[[1]]) %>% # combine vector contamination for sample (contamout.tibble) and positives (contam.tibble)
        as.data.frame() %>%
        ggplot(aes(x = vector_contamination)) + 
        geom_histogram() # y is the count of ASVs
    
      ggsave(
        plot = vc_plot_1,
        here(config_file$output_path,config_file$region,"Output_plots","step2_Vector_Contamination_Barcode_1.png"),
        device = "png",
        width = 12,
        height = 8,
        units = "in"
      )

} else { # If there isn't a positive control output message
  message("Positive control not found or didn't pass filters — skipping analysis")
}

# if you want to look at output:
# ASV_nested$`positive control`

```

### 2.3: Subtract the composition of the positive controls from the environment samples

- This is done ASV by ASV
- This could be translated to take the highest rate of index hopping observed for any non-positive control ASV and applied across all ASVs

```{r subtract-pos-from-env, echo=FALSE, warning=FALSE, message=FALSE}
# Only run tag jumping in and out of positive controls if positive control in samples:
if (("positive control" %in% merged_table_long_v2$samp_category) &&
     ("pass" %in% merged_table_long_v2$Pos_pass_filters) | 
     config_file$region=="18Sv4" && exists("OME_18Sv4_pos")) {
  message("Positive control found & passed filters — running code")
  if(config_file$region=="18Sv4") {
    max_vector_contamination <- OME_18Sv4_pos %>%
      mutate(vector_contamination=max_contam)
  } else {
  ## Find out what is the top contaminant across samples (not including known positive ASVs)
  max_vector_contamination <- rbind(ASV_nested$contam_tibble[[1]], # contaminant proportions in positive controls
                                    ASV_nested$contamout_tibble[[1]]) %>% # # contaminant proportions in samples
    left_join(hash_key) %>% # add taxonomy from taxonomy table
    # remove known positive control species and genus:
    filter(., Taxon != config_file$step2_positive_species_taxonomic_path & # with genus
                 !!sym(config_file$step2_positive_lowest_taxonomic_lvl) != config_file$step2_positive_lowest_taxonomic_lvl_id 
           & Domain!="Unknown") %>% # with higher level taxonomy
    # include target taxonomy here
    arrange(desc(vector_contamination)) %>%
    dplyr::select(-c(Domain:Species)) %>%
    dplyr::slice(1) 
  }
  ## Add to ASV_nested
  ASV_nested <- ASV_nested %>%
      ungroup() %>%
      mutate(Step2_cleaned_tibble = purrr::map(sample, function(.x) {
        .x %>%
          dplyr::group_by(sample) %>%
          mutate(TotalReadsperSample = sum(reads)) %>% # calculate total reads per sample
          mutate(Updated_nReads = case_when(
            is.na(max_vector_contamination$vector_contamination[[1]]) ~ reads,
            !is.na(max_vector_contamination$vector_contamination[[1]]) ~ reads-( 
              ceiling(max_vector_contamination$vector_contamination[[1]] * TotalReadsperSample)
          ))) %>%
          filter(Updated_nReads > 0) %>%
          ungroup() %>%
          dplyr::select(sample, seq_id, reads=Updated_nReads) # reads, Updated_nReads, TotalReadsperSample)
      }))
    
    ASVs_retained_step2 <- ASV_nested$Step2_cleaned_tibble[[1]]$seq_id %>% unique()
  # which species were removed?: 
 ASV_nested$sample[[1]] %>%
   filter(., !seq_id %in% ASVs_retained_step2) %>%
   group_by(seq_id, Taxon) %>%
   dplyr::summarise(n(), sum_reads =sum(reads) )

  #### create summary file (Summary_file_2) 
  ASV_summary <- ASV_nested %>%
    dplyr::mutate(Miseq_region, Summary_2 = purrr::map(Step2_cleaned_tibble, ~ how.many(ASVtable = ., round = "1_Jump")), .keep = "none") %>%
    left_join(ASV_summary) %>% # use left join when there are many miseq runs to join
    mutate(Summary = map2(Summary, Summary_2, bind_rows)) %>%
    dplyr::select(-Summary_2)
  
  ASV_summary$Summary
  
  } else {
  message("Positive control not found or didn't pass filters - skipping analysis")
     ASV_nested <- ASV_nested %>%
      ungroup() %>%
      mutate(Step2_cleaned_tibble = purrr::map(sample, function(.x) {
      .x %>%
        dplyr::select(sample, seq_id, reads)
      }))
     
  ## Summary file so we have this step in our dataframe:
  ASV_summary <- ASV_nested %>%
    dplyr::mutate(Miseq_region, Summary_2 = purrr::map(Step2_cleaned_tibble, ~ how.many(ASVtable = ., round = "1_Jump")), .keep = "none") %>%
    left_join(ASV_summary) %>% # use left join when there are many miseq runs to join
    mutate(Summary = map2(Summary, Summary_2, bind_rows)) %>%
    dplyr::select(-Summary_2)
  
  ASV_summary$Summary
  
  }

```

## Step 3: Remove negative control contaminants

### Step 3.1: Investigate contaminants found in negative controls.

```{r investigate-neg-controls, echo=FALSE, warning=FALSE, message=FALSE}
# Only run negative control analysis if there's a negative control in samples:
if ("negative control" %in% merged_table_long_v2$samp_category) {
  message("Negative control found — running code")

  ## load in negative controls as separate df:
  neg_controls <- merged_table_long_v2 %>% # create dataframe of only samples that are negative controls.
    filter(samp_category=="negative control") # only blanks (negative controls)

  ## Generate Long Data Frame of Negative Controls
  negative_long <- ASV_nested$`negative control`[[1]] %>% # save negative control long format data frame in nested table
    filter(Not_off_target==TRUE & neg_total_reads >= 1000) 
  
  ## Save sequences & samples that are off target
  write_csv(ASV_nested$`negative control`[[1]] %>% filter(Not_off_target==FALSE & neg_total_reads < 1000),
            here(config_file$output_path,config_file$region,"Output_csv","step2_nc_filtered_out.csv"))  
  
  ### Reads per Contaminant:
  negative_reads <- negative_long %>%
      group_by(seq_id) %>%
      dplyr::summarise(sum_reads_neg = sum(reads)) %>% # summarise total reads per ASV in blanks
      left_join(hash_key) %>% # join with taxonomy (to see taxa of contaminants)
      arrange(desc(sum_reads_neg)) # order ASVs with highest counts first
  
  negative_reads

} else {
  message("Negative control not found - skipping analysis")
}

```

### Step 3.3: Which species are in the blanks and what are their occurrences & read counts

Calculate across all sequence id's

```{r id-neg-control-sp, echo=FALSE, warning=FALSE, message=FALSE}
# Only run negative control analysis if there's a negative control in samples:
if ("negative control" %in% merged_table_long_v2$samp_category) {
  message("Negative control found — running code") 
    # if there are associated samples
    if(!is.null(step3_associated_nc)) {
       message("Using associated negatives only") 
        ### Generate Read Counts across all samples and negative controls:
        read_counts_step3 <- ASV_nested %>%
            dplyr::select(Step2_cleaned_tibble) %>% # pos contaminants removed tibble
            unnest(Step2_cleaned_tibble) %>% # This stage is a long table
            ungroup() %>% # unnest tibble to work with
            group_by(seq_id, sample) %>% 
            arrange(desc(reads))
        ### Generate Occurrence Counts
        number_of_occurrences <-ASV_nested %>%
              dplyr::select(Step2_cleaned_tibble) %>%
              unnest(Step2_cleaned_tibble) %>% # This stage is a long table
              ungroup() %>%
              group_by(seq_id, sample) %>%
              dplyr::count() # Count number of times ASVs are found across all samples
        ### look at ASVs only found in negative controls (and not samples):
        setdiff(negative_long$seq_id, number_of_occurrences$seq_id)
          
        ### look at taxa shared between ASVs found in negative control and samples:
        intersect(negative_long$seq_id, number_of_occurrences$seq_id) %>% # which ASVs are shared?
            as_tibble() %>%
            dplyr::rename(seq_id = value) %>%
            left_join(hash_key) # add taxa to ASVs 
        
        ## Create column of whether or not an ASV was found in blanks or samples
        number_of_occurrences <- number_of_occurrences %>%
            mutate(Contam = if_else(seq_id %in% negative_long$seq_id, "In Negatives", "In Samples Only")) %>%
            arrange(desc(n))
          
    } else {
       message("Using all negatives") 
        ### Generate Read Counts across all samples and negative controls:
        read_counts_step3 <- ASV_nested %>%
            dplyr::select(Step2_cleaned_tibble) %>% # pos contaminants removed tibble
            unnest(Step2_cleaned_tibble) %>% # This stage is a long table
            ungroup() %>% # unnest tibble to work with
            group_by(seq_id, sample) %>% 
            dplyr::summarise( # summarize read counts per ASV
              sum_reads = sum(reads), # Total reads per ASV
              mean_reads = mean(reads), # Mean reads per ASV
              median_reads = median(reads), # Median reads per ASV
              max_reads = max(reads), # Max reads per ASV
              sd_reads = sd(reads), # SD reads per ASV
              min_reads = min(reads) # Minimum reads per ASV
              ) %>%
            arrange(desc(sum_reads))
         ### Generate Occurrence Counts
          number_of_occurrences <-ASV_nested %>%
              dplyr::select(Step2_cleaned_tibble) %>%
              unnest(Step2_cleaned_tibble) %>% # This stage is a long table
              ungroup() %>%
              group_by(seq_id) %>%
              dplyr::count() # Count number of times ASVs are found across all samples
          ### look at ASVs only found in negative controls (and not samples):
          setdiff(negative_long$seq_id, number_of_occurrences$seq_id)
          
          ### look at taxa shared between ASVs found in negative control and samples:
          intersect(negative_long$seq_id, number_of_occurrences$seq_id) %>% # which ASVs are shared?
            as_tibble() %>%
            dplyr::rename(seq_id = value) %>%
            left_join(hash_key) # add taxa to ASVs 
        
          ## Create column of whether or not an ASV was found in blanks or samples
          number_of_occurrences <- number_of_occurrences %>%
            mutate(Contam = if_else(seq_id %in% negative_long$seq_id, "In Negatives", "In Samples Only")) %>%
            arrange(desc(n))
    }
  } else {
  message("Negative control not found")
  }

```

Blanks and negative controls

```{r neg-con-sp-samples, echo=FALSE, warning=FALSE, message=FALSE}
# Only run negative control analysis if there's a negative control in samples:
if ("negative control" %in% merged_table_long_v2$samp_category) {
  message("Negative control found — running code")
  
  ### Which Species are in the blanks and what are their occurrences & read counts
   neg_comparison <- number_of_occurrences %>%
    filter(., Contam == "In Negatives") %>% # only look at ASV occurrences found in negative controls
    left_join(hash_key) %>% # merge ASV occurrence count with taxonomy
    left_join(negative_reads) %>% # join occurence with sum of negative reads per ASV
    left_join(read_counts_step3) %>% # join with summary of each ASV across all samples
    mutate(., max_n = max(number_of_occurrences$n))

  ### How would removing these ASVs change the number of results?
  ASV_nested %>%
    dplyr::select(Step2_cleaned_tibble) %>%
    unnest(Step2_cleaned_tibble) %>% # pull out previous filter step counts
    filter(., seq_id %in% neg_comparison$seq_id) %>% # select only ASVs in negative controls
    group_by(seq_id) %>%
    dplyr::summarise(
      tot_reads_with_negASVs = sum(reads), # summarize reads for ASVs in neg control
      mean_reads_with_negASVs = round(mean(reads),digits=0), # mean of reads for ASVs in neg control
      occurrences_with_negASVs = n_distinct(sample) # count how many times ASVs are found across all samples
    )

  ASV_nested %>%
    dplyr::select(Step2_cleaned_tibble) %>%
    unnest(Step2_cleaned_tibble) %>% # pull out previous filter step counts
    filter(., !seq_id %in% neg_comparison$seq_id) %>% # select only ASVs not in negative controls
    group_by(seq_id) %>%
    dplyr::summarise(
      tot_reads_without_negASVs = sum(reads), # summarize reads for ASVs not found in blanks
      mean_reads_without_negASVs = round(mean(reads),digits=0),  # mean of reads for ASVs not found in blanks
      occurrences_without_negASVs = n_distinct(sample) # count how many times ASVs are found across all samples
    )

} else {
  message("Negative control not found - skipping analysis")
}

```

Plot of negative controls counts

```{r plot-neg-con, echo=FALSE, warning=FALSE, message=FALSE}
# Only run negative control analysis if there's a negative control in samples:
if ("negative control" %in% merged_table_long_v2$samp_category) {
  message("Negative control found — running code")
  
  ### Plot Occurrences
  plot_occ <- number_of_occurrences %>% # plotting ASV occurrence across samples
    group_by(seq_id, Contam) %>%
    summarise(n=sum(n)) %>% # summarise by sequence not reads
    filter(., n > 3) %>% # only plot ASVs that occur in at least 6 samples
    ggplot(., aes(x = reorder(seq_id, n), y = n, fill = Contam)) + # fill with Contam status (in neg control or only in sample)
    geom_bar(
      stat =
        "identity"
    ) +
    facet_wrap(~Contam)

  ggsave(
    plot = plot_occ,
    here(config_file$output_path,config_file$region,"Output_plots",
         "step3_ASVs_from_Negative_Control_Occurrence_Plot.png"),
    device = "png",
    width = 12,
    height = 8,
    units = "in"
  )

} else {
  message("Negative control not found - skipping analysis")
}

```

Create negative control dataframes for filtering

```{r format-neg-control-df, echo=FALSE, warning=FALSE, message=FALSE}
##### Identify blank samples
if (!is.null(ASV_nested$'negative control'[[1]]) && nrow(negative_long) > 0) { # only run if negative controls in sample
  message("Negative control found — running code")
  # if there are associated samples
    if(!is.null(step3_associated_nc)) {
       message("Using associated negatives only")
      # save associated column name to filter by:
      assoc_id_column<-config_file$step3_assoc_nc_field
      # make df of samples and asvs for later use:  
      blank_samples_asvs <- negative_long %>% # Sample names and ASVs for blanks
        dplyr::select(sample, seq_id, reads, !!sym(assoc_id_column)) # only keep selected columns
      # create an otu table to match with samples below for decontam:
      blank_ASVs_1 <- blank_samples_asvs %>%
        dplyr::select(seq_id, reads, sample) %>% # select only necessary columns
        
        pivot_wider( # pivot wider so read counts are now below each sample name
          names_from = sample,
          values_from = reads,
          values_fill = 0
        )
      # blank lookup table for matching associated samples:
      blank_lookup <- negative_long %>% dplyr::select(sample, !!sym(assoc_id_column)) %>%
       separate_longer_delim(!!sym(assoc_id_column), delim = regex("[,]+")) %>%
        # Group by the specific sample ID
        group_by(!!sym(assoc_id_column)) %>%
        # Consolidate the negative control names (sample column) 
        # Using unique() ensures you don't list the same blank twice for one sample
        summarize(sample = paste(unique(sample), collapse = ","), .groups = "drop") %>%
        # Flip it back: Now group by the Blanks to get the format microDecon needs
        group_by(sample) %>%
        summarize(!!sym(assoc_id_column) := 
                    paste(!!sym(assoc_id_column), collapse = ","), .groups = "drop") %>%
        mutate(!!sym(assoc_id_column) := na_if(!!sym(assoc_id_column), "NA"))
     
      #### Columns to Ignore for Grouping
      blank_samples <- negative_long$sample %>% unique()
    } else {
      message("No associated samples - using all negative controls")
      blank_samples_asvs <- negative_long %>% # Sample names and ASVs for blanks
        dplyr::select(sample, seq_id, reads) # only keep selected columns
      # create an otu table to match with samples below for decontam:
      blank_ASVs_1 <- blank_samples_asvs %>%
        dplyr::select(seq_id, reads, sample) %>% # select only necessary columns
        pivot_wider( # pivot wider so read counts are now below each sample name
          names_from = sample,
          values_from = reads,
          values_fill = list(reads = 0)
        )
      
     
      #### Columns to Ignore for Grouping
      blank_samples <- ASV_nested$'negative control'[[1]]$sample %>% unique()
    }
} else { # If there are no blanks in samples, skip this step.
  message("Negative control not found")
  blank_samples <- ASV_nested$'negative control'[[1]]$sample %>% unique()
}
```

### Step 3.4 option A: Format and run for microDecon

Format dataframe for (microDecontam)[https://github.com/donaldtmcknight/microDecon]

```{r nc-microdecon-formatting, echo=FALSE, warning=FALSE, message=FALSE}
# Only run if negative control found:
if (!is.null(ASV_nested$'negative control'[[1]]) && nrow(negative_long) > 0) {
  message("Negative control found — running code")
  ##### Save previous steps to use for microDecon
  step_2 <- ASV_nested$Step2_cleaned_tibble[[1]]
  #### Pivot to Wide format
  step_2_wide <- step_2 %>%
    dplyr::select(seq_id, reads, sample) %>% # select only necessary columns
    pivot_wider( # pivot wider so read counts are now below each sample name
      names_from = sample,
      values_from = reads,
      values_fill = list(reads = 0)
    )

  #### Arrange Blank Samples to be the first 3 columns
  step_2_wide_1 <- step_2_wide %>%
    full_join(blank_ASVs_1, ., by="seq_id") %>% # add blanks to df
    # replace any values found in blank but not samples with zero (or vice versa)
    mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
    left_join(hash_key %>% select("seq_id","Taxon")) %>% # add taxonomy
    relocate(Taxon, .after = last_col()) %>% # move taxonomy to the end of the dataframe
    relocate(blank_samples, .after = seq_id) %>% # move blanks samples to the first three columns after the ASV column
    as.data.frame()

} else {
  message("Negative control not found - skipping analysis")
   ##### Save previous steps to use for microDecon
  step_2 <- ASV_nested$Step2_cleaned_tibble[[1]]
}

```

Run microdecontam with associated samples

```{r nc-microdecon-assoc-samples, echo=FALSE, warning=FALSE, message=FALSE}
# Only run if negative control found:
if (!is.null(ASV_nested$'negative control'[[1]]) && 
    nrow(negative_long) > 0 &&
    config_file$step3_nc_filt=="microdecontam" &&
    !is.null(step3_associated_nc)) {
      message("Running Microdecontam with associated samples")
        # 1. Initialize a list to store results
        decon_list <- list()
        
        for (i in 1:nrow(blank_lookup)) {
          # Identify the specific blank and its associated samples for this iteration
          raw_blanks <- unlist(strsplit(blank_lookup$sample[i], "[,]+"))
          current_blanks <- raw_blanks[raw_blanks %in% colnames(step_2_wide_1)]
          # Split the string and keep ONLY those that actually exist in step.2_wide_1
          raw_assocs <- unlist(strsplit(as.character(blank_lookup[[assoc_id_column]][i]), "[,]+"))
          current_assocs <- raw_assocs[raw_assocs %in% colnames(step_2_wide_1)]
          
          # 3. Validation Check
          # Only run if the blank exists AND there is at least one sample to clean
          if (length(current_blanks) > 0 && length(current_assocs) > 0) {
            
              # 2. Subset step.2_wide_1 for ONLY this group
              # microDecon order: [OTU IDs (col 1), Blanks, Samples, Taxon (last)]
              iteration_data <- step_2_wide_1 %>%
                select(seq_id, 
                       all_of(current_blanks), 
                       all_of(current_assocs), 
                       Taxon)
            
              # 3. Calculate sample_counter for this specific subset
              # We look at the metadata for only the samples in 'current_assocs'
              sub_sample_counter <- data.frame(sample = current_assocs) %>%
                left_join(merged_table_long_v2, by = "sample") %>%
                select(sample, samp_category) %>% 
                unique() %>%
                group_by(samp_category) %>%
                dplyr::count()
              
              # 4. Run decon for this group
              temp_decon <- decon(
                data = iteration_data,
                numb.blanks = length(current_blanks),
                numb.ind = sub_sample_counter$n,
                taxa = TRUE,
                runs = config_file$step3_runs,
                thresh = config_file$step3_thresh,
                prop.thresh = config_file$step3_prop_thresh,
                regression = config_file$step3_regression
              )
              
              # 5. Save the decontaminated table
              # We extract $decon.table and store it in our list
              decon_list[[i]] <- temp_decon$decon.table
            } else {
                message(paste("Skipping row", i, ": Blank or Samples not found in data."))
              }
        }
        
        if (length(decon_list)==0 && is.na(blank_lookup[[assoc_id_column]])) {
          message(paste("Skipping microdecontam, no associated samples"))
          no_associated_samples<-TRUE # use this later, to make sure workflow isn't interrupted by no negative control filtering.
        } else {
        # 6. Merge all results back together
        # Since each iteration contains the 'Taxon' and 'seq_id' columns, 
        # we should join them by those IDs to get one master wide table.
        step_3_decon <- purrr::reduce(decon_list, full_join, by = c("seq_id", "Taxon")) %>%
          select(-contains("Mean.blank")) # remove averaged blanks from microdecon output.
        
        ### Convert to Long Data
        columns <- colnames(step_3_decon)
        remove <- c("seq_id", "Taxon")
        # columns to gather in pivot longer
        gathercols <- columns[!columns %in% remove]
      
        # Save deconamtinated dataframe
        step_3_decon_clean <-
          step_3_decon %>%
          pivot_longer(.,
            cols = gathercols,
            names_to = "sample",
            values_to = "reads"
          )
        # save rds for later just in case:
        saveRDS(step_3_decon, here(config_file$output_path,config_file$region,"Output_R","step_3_decon.RDS"))
        }
    } else {
    message("Either no negative controls, not running decontam analysis, or no associated sample list")
    }

```

Run microdecontam across all samples if no associated samples provided

```{r nc-microdecon-no-assoc-samples, echo=FALSE, warning=FALSE, message=FALSE}  
if (!is.null(ASV_nested$'negative control'[[1]]) && 
    nrow(negative_long) > 0 &&
    config_file$step3_nc_filt=="microdecontam" &&
    is.null(step3_associated_nc)) {
        message("Running Microdecontam across all samples")
        # Number of 'negative control'
        blank_num <- ASV_nested$'negative control'[[1]]$sample %>%
          unique() %>% 
          length()
        # columns to remove 
        col_delete <- append(blank_samples, c("Taxon", "seq_id"))
      
        # count number of samples
        sample_counter <- colnames(step_2_wide_1) %>%
            as.data.frame() %>%
            filter(., !(. %in% col_delete)) %>%
            dplyr::rename(sample = '.') %>%
            left_join(merged_table_long_v2, by = "sample") %>%
            select("sample","samp_category") %>% unique() %>%
            group_by(samp_category) %>%
            dplyr::count()
      
        step_3_decon <- decon( # run microdecontam with settings from yaml:
          data = step_2_wide_1,
          numb.blanks = blank_num,
          numb.ind = sample_counter$n,
          taxa = T,
          runs = config_file$step3_runs,
          thresh = config_file$step3_thresh,
          prop.thresh = config_file$step3_prop_thresh,
          regression = config_file$step3_regression
        )
        
        ### Convert to Long Data
        columns <- colnames(step_3_decon$decon.table)
        remove <- c("seq_id", "Taxon")
        # columns to gather in pivot longer
        gathercols <- columns[!columns %in% remove]
      
        # Save deconamtinated dataframe
        step_3_decon_clean <-
          step_3_decon$decon.table %>%
          pivot_longer(.,
            cols = gathercols,
            names_to = "sample",
            values_to = "reads"
          )
      # save rds for later just in case
      saveRDS(step_3_decon, here(config_file$output_path,config_file$region,"Output_R","step_3_decon.RDS"))
      
  } else {
    message("Either no negative controls, not running decontam analysis")
    }
```   

Final Cleanup: Add to ASV_nested and ASV_summary

```{r nc-microdecon-final-cleanup, echo=FALSE, warning=FALSE, message=FALSE}
if (!is.null(ASV_nested$'negative control'[[1]]) && 
    nrow(negative_long) > 0 &&
    config_file$step3_nc_filt=="microdecontam" && 
    !exists("no_associated_samples")) {
        message("Running Microdecontam across all samples")

    # format decontam output to create dataframe
    step_3_decon_clean_tibble <- 
      step_3_decon_clean %>%
        as_tibble() %>%
        dplyr::select(-Taxon) %>%
        filter(., !sample %in% c("Mean.blank", "fake_asv_1")) %>% # remove the fake blanks if they were added above
        filter(., !seq_id == "fake_asv_1") %>% # remove the fake ASV if they were added above
        filter(., reads > 0) %>%
        filter(., !sample %in% blank_samples)
    
    ## Add to ASV_nested
    ASV_nested <- ASV_nested %>%
      mutate(., Step3_tibble = list(step_3_decon_clean_tibble))
    
    ## Summary_file_3
    ASV_summary <- ASV_nested %>%
      transmute(Miseq_region, Summary_3 = purrr::map(Step3_tibble, ~ how.many(ASVtable = ., round = "3_microDecon"))) %>%
      left_join(ASV_summary) %>%
      mutate(Summary = map2(Summary, Summary_3, bind_rows)) %>%
      dplyr::select(-Summary_3)
    
    ASV_summary$Summary
    
  } else {
    message("Either no negative controls, or not running decontam analysis")
    }

```

### Step 3.4 option B: Filter negative controls by proportion

- If more than one negative control associated with samples, get the mean proportions and use that for each seq_id in each sample

```{r nc-prop-assoc-samples, echo=FALSE, warning=FALSE, message=FALSE}
# Only run if negative control found:
if (!is.null(ASV_nested$'negative control'[[1]]) && 
    nrow(negative_long) > 0 &&
    config_file$step3_nc_filt=="proportion" &&
    !is.null(step3_associated_nc)) {
  message("Running Proportional filtering with associated samples")
        # 1. Initialize a list to store results
        prop_filt_list <- list()
        
        for (i in 1:nrow(blank_lookup)) {
          # Identify the specific blank and its associated samples for this iteration
          raw_blanks <- unlist(strsplit(blank_lookup$sample[i], "[,]+"))
          current_blanks <- raw_blanks[raw_blanks %in% colnames(step_2_wide_1)]
          # Split the string and keep ONLY those that actually exist in step.2_wide_1
          raw_assocs <- unlist(strsplit(as.character(blank_lookup[[assoc_id_column]][i]), "[,]+"))
          current_assocs <- raw_assocs[raw_assocs %in% colnames(step_2_wide_1)]
          
          # 3. Validation Check
          # Only run if the blank exists AND there is at least one sample to clean
          if (length(current_blanks) > 0 && length(current_assocs) > 0) {
            
              # 2. Subset step_2_wide_1 for negative controls and calculate proportion
              blank_prop <- step_2_wide_1 %>% 
                select(seq_id,all_of(current_blanks), Taxon) %>%
                mutate(across(where(is.numeric), ~ .x / sum(.x, na.rm = TRUE))) %>%
                mutate(blank_mean=rowMeans(across(where(is.numeric) & !seq_id), na.rm = TRUE)) %>%
                select(seq_id,blank_mean)
              
              
              # 3. Use blank_prop to subtract from associated samples
                # save current reads dataframe:
                assoc_reads<-step_2_wide_1 %>% 
                  select(seq_id, all_of(current_assocs), Taxon)
                # save total reads:
                assoc_reads_tot<-assoc_reads %>%
                  summarise(across(current_assocs, sum, na.rm = TRUE))

                # calculate proportion in each sample and subtract negative control prop
                filtered_samples <- assoc_reads %>%
                  mutate(across(where(is.numeric), ~ .x / sum(.x, na.rm = TRUE))) %>%
                  left_join(blank_prop, by="seq_id") %>%
                  # subtract blank proportions for each seq_id from each sample
                  mutate(across(where(is.numeric) & !seq_id & !blank_mean, ~ . - blank_mean)) %>% 
                  select(-"blank_mean") %>% # remove blank_mean
                  # replace negatives with 0
                  mutate(across(where(is.numeric) & !seq_id, ~ ifelse(. < 0, 0, .))) %>%
                  # multiply each proportion by total reads to add back to final dataframe:
                  mutate(across(where(is.numeric), ~ round(.x * assoc_reads_tot[[cur_column()]][1], digits=0)))

              
                # 4. Save the prop filtered table
                prop_filt_list[[i]] <- filtered_samples
              } else {
                  message(paste("Skipping row", i, ": Blank or Samples not found in data."))
                }
        }
        if (length(prop_filt_list)==0 && is.na(blank_lookup[[assoc_id_column]])) {
          message(paste("Skipping microdecontam, no associated samples"))
          no_associated_samples<-TRUE # use this later, to make sure workflow isn't interrupted by no negative control filtering.
        } else {
        # 5. Merge all results back together
        # Since each iteration contains the 'Taxon' and 'seq_id' columns, 
        # we should join them by those IDs to get one master wide table.
        step_3_prop <- purrr::reduce(prop_filt_list, full_join, by = c("seq_id", "Taxon"))
        
        ### Convert to Long Data
        columns <- colnames(step_3_prop)
        remove <- c("seq_id", "Taxon")
        # columns to gather in pivot longer
        gathercols <- columns[!columns %in% remove]
      
        # Save deconamtinated dataframe
        step_3_prop_clean <-
          step_3_prop %>%
          pivot_longer(.,
            cols = gathercols,
            names_to = "sample",
            values_to = "reads"
          )
      # save rds for later just in case:
      saveRDS(step_3_prop, here(config_file$output_path,config_file$region,"Output_R","step_3_prop.RDS"))
        
      } 
    } else {
    message("Either no negative controls, not running proportion filtering, or no associated sample list")
    }

```

Run proportion filtering across all samples if no associated samples provided

```{r nc-prop-no-assoc-samples, echo=FALSE, warning=FALSE, message=FALSE}  
if (!is.null(ASV_nested$'negative control'[[1]]) && 
    nrow(negative_long) > 0 &&
    config_file$step3_nc_filt=="proportion" &&
    is.null(step3_associated_nc)) {
        message("Running proportion filtering across all samples")
          # Identify the blanks
          all_blanks <- unlist(strsplit(blank_lookup$sample, "[,]+"))
  
          # 2. Subset step_2_wide_1 for negative controls and calculate proportion
          blank_prop <- step_2_wide_1 %>% 
            select(seq_id,all_of(current_blanks), Taxon) %>%
            mutate(across(where(is.numeric), ~ .x / sum(.x, na.rm = TRUE))) %>%
            mutate(blank_mean=rowMeans(across(where(is.numeric) & !seq_id), na.rm = TRUE)) %>%
            select(seq_id,blank_mean)
              
          # 3. Use blank_prop to subtract from associated samples
          # save current reads dataframe:
          all_reads<-step_2_wide_1 %>% 
            select(-any_of(all_blanks)) # remove blanks from dataframe
          # save total reads:
          all_reads_tot<-all_reads %>%
            summarise(across(where(is.numeric) & !seq_id, sum, na.rm = TRUE)) 

          # calculate proportion in each sample and subtract negative control prop
          step_3_prop <- all_reads %>%
            mutate(across(where(is.numeric) & !seq_id, ~ .x / sum(.x, na.rm = TRUE))) %>%
            left_join(blank_prop, by="seq_id") %>%
            # subtract blank proportions for each seq_id from each sample
            mutate(across(where(is.numeric) & !seq_id & !blank_mean, ~ . - blank_mean)) %>% 
             select(-"blank_mean") %>% # remove blank_mean
             # replace negatives with 0
             mutate(across(where(is.numeric) & !seq_id, ~ ifelse(. < 0, 0, .))) %>%
             # multiply each proportion by total reads to add back to final dataframe:
             mutate(across(where(is.numeric), ~ round(.x * all_reads_tot[[cur_column()]][1], digits=0)))
        
        ### Convert to Long Data
        columns <- colnames(step_3_prop)
        remove <- c("seq_id", "Taxon")
        # columns to gather in pivot longer
        gathercols <- columns[!columns %in% remove]
      
        # Save deconamtinated dataframe
        step_3_prop_clean <-
          step_3_prop %>%
          pivot_longer(.,
            cols = gathercols,
            names_to = "sample",
            values_to = "reads"
          )
      # save rds for later just in case:
      saveRDS(step_3_prop, here(config_file$output_path,config_file$region,"Output_R","step_3_prop.RDS"))
      
  } else {
    message("Either no negative controls, not running proportion filtering")
    }
```   

Final Cleanup: Add to ASV_nested and ASV_summary

```{r nc-prop-final-cleanup, echo=FALSE, warning=FALSE, message=FALSE}
if (!is.null(ASV_nested$'negative control'[[1]]) && 
    nrow(negative_long) > 0 &&
    config_file$step3_nc_filt=="proportion" &&
    !exists("no_associated_samples")) {
        message("Running Microdecontam across all samples")

    # format prop filtered output to create dataframe
    step_3_prop_clean_tibble <- 
      step_3_prop_clean %>%
        as_tibble() %>%
        dplyr::select(-Taxon) %>%
        filter(., reads > 0) %>%
        filter(., !sample %in% blank_samples)
    
    ## Add to ASV_nested
    ASV_nested <- ASV_nested %>%
      mutate(., Step3_tibble = list(step_3_prop_clean_tibble))
    
    ## Summary_file_3
    ASV_summary <- ASV_nested %>%
      transmute(Miseq_region, Summary_3 = purrr::map(Step3_tibble, ~ how.many(ASVtable = ., round = "3_NC_prop.filt"))) %>%
      left_join(ASV_summary) %>%
      mutate(Summary = map2(Summary, Summary_3, bind_rows)) %>%
      dplyr::select(-Summary_3)
    
    ASV_summary$Summary

  } else {
    message("Either no negative controls, not running proportion filtering")
    }

```

### Step 3.4 option C: Filter any ASVs found in negative controls

- If more than one negative control associated with samples, remove any asvs that showed up in any blanks

```{r nc-all-asvs-associated, echo=FALSE, warning=FALSE, message=FALSE}
# Only run if negative control found:
if (!is.null(ASV_nested$'negative control'[[1]]) && 
    nrow(negative_long) > 0 &&
    config_file$step3_nc_filt=="occurrence" &&
    !is.null(step3_associated_nc)) {
  message("Running occurrence filtering with associated samples")
        # 1. Initialize a list to store results
        occ_filt_list <- list()
        
        for (i in 1:nrow(blank_lookup)) {
          # Identify the specific blank and its associated samples for this iteration
          raw_blanks <- unlist(strsplit(blank_lookup$sample[i], "[,]+"))
          current_blanks <- raw_blanks[raw_blanks %in% colnames(step_2_wide_1)]
          # Split the string and keep ONLY those that actually exist in step_2_wide_1
          raw_assocs <- unlist(strsplit(as.character(blank_lookup[[assoc_id_column]][i]), "[,]+"))
          current_assocs <- raw_assocs[raw_assocs %in% colnames(step_2_wide_1)]
          
          # 3. Validation Check
          # Only run if the blank exists AND there is at least one sample to clean
          if (length(current_blanks) > 0 && length(current_assocs) > 0) {
            
              # 2. Subset step_2_wide_1 for negative controls and calculate proportion
              blank_occ <- step_2_wide_1 %>% 
                select(seq_id,all_of(current_blanks), Taxon) %>%
                # keep only ASVs with sums greater than zero across all samples
                filter(rowSums(across(where(is.numeric)), na.rm = TRUE) > 0) %>%
                pivot_longer(.,cols = -c("seq_id", "Taxon"),
                             names_to = "sample", values_to = "reads") %>%
                ungroup() %>%
                group_by(seq_id, sample, Taxon) %>%
                dplyr::count() # count occurence of each ASV
              
              # 3. Use blank_occ to subtract any asvs found in blanks:
              # calculate proportion in each sample and subtract negative control prop
              filtered_occ_samples <- step_2_wide_1 %>% 
                select(seq_id, all_of(current_assocs), Taxon) %>%
                filter(., !seq_id %in% blank_occ$seq_id)
            
              # 4. Save the prop filtered table
              occ_filt_list[[i]] <- filtered_occ_samples
              } else {
                message(paste("Skipping row", i, ": Blank or Samples not found in data."))
                }
        }
        if (length(occ_filt_list)==0 && is.na(blank_lookup[[assoc_id_column]])) {
          message(paste("Skipping microdecontam, no associated samples"))
          no_associated_samples<-TRUE # use this later, to make sure workflow isn't interrupted by no negative control filtering.
        } else {
        # 5. Merge all results back together
        # Since each iteration contains the 'Taxon' and 'seq_id' columns, 
        # we should join them by those IDs to get one master wide table.
        step_3_occ <- purrr::reduce(occ_filt_list, full_join, by = c("seq_id", "Taxon"))
        
        ### Convert to Long Data
        columns <- colnames(step_3_occ)
        remove <- c("seq_id", "Taxon")
        # columns to gather in pivot longer
        gathercols <- columns[!columns %in% remove]
      
        # Save deconamtinated dataframe
        step_3_occ_clean <-
          step_3_occ %>%
          pivot_longer(.,
            cols = gathercols,
            names_to = "sample",
            values_to = "reads"
          )
      # save rds for later just in case:
      saveRDS(step_3_occ, here(config_file$output_path,config_file$region,"Output_R","step_3_occ.RDS"))
        
    } 
    } else {
    message("Either no negative controls, not running occurrence filtering, or no associated sample list")
    }

```

Run proportion filtering across all samples if no associated samples provided

```{r nc-all-asvs-no-associated, echo=FALSE, warning=FALSE, message=FALSE}  
if (!is.null(ASV_nested$'negative control'[[1]]) && 
    nrow(negative_long) > 0 &&
    config_file$step3_nc_filt=="occurrence" &&
    is.null(step3_associated_nc)) {
        message("Running occurrence filtering across all samples")
          # Identify the blanks
          all_blanks <- unlist(strsplit(blank_lookup$sample, "[,]+"))
          
            # 2. Subset step_2_wide_1 for negative controls and calculate proportion
            blank_occ <- step_2_wide_1 %>% 
              select(seq_id,all_of(all_blanks), Taxon) %>%
              # keep only ASVs with sums greater than zero across all samples
              filter(rowSums(across(where(is.numeric)), na.rm = TRUE) > 0) %>%
              pivot_longer(.,cols = -c("seq_id", "Taxon"),
                             names_to = "sample", values_to = "reads") %>%
              ungroup() %>%
              group_by(seq_id, sample, Taxon) %>%
              dplyr::count() # count occurence of each ASV
              
            # 3. Use blank_occ to subtract any asvs found in blanks:
            # calculate proportion in each sample and subtract negative control prop
            step_3_occ <- step_2_wide_1 %>% 
              select(-any_of(all_blanks)) %>%# remove blanks from dataframe
              filter(., !seq_id %in% blank_occ$seq_id)
  
        ### Convert to Long Data
        columns <- colnames(step_3_occ)
        remove <- c("seq_id", "Taxon")
        # columns to gather in pivot longer
        gathercols <- columns[!columns %in% remove]
      
        # Save deconamtinated dataframe
        step_3_occ_clean <-
          step_3_occ %>%
          pivot_longer(.,
            cols = gathercols,
            names_to = "sample",
            values_to = "reads"
          )
      # save rds for later just in case:
      saveRDS(step_3_occ, here(config_file$output_path,config_file$region,"Output_R","step_3_occ.RDS"))
      
  } else {
    message("Either no negative controls, not running occurrence filtering")
    }
```   

Final Cleanup: Add to ASV_nested and ASV_summary

```{r nc-all-asvs-cleanup, echo=FALSE, warning=FALSE, message=FALSE}
if (!is.null(ASV_nested$'negative control'[[1]]) && 
    nrow(negative_long) > 0 &&
    config_file$step3_nc_filt=="occurrence" &&
    !exists("no_associated_samples")) {
        message("Running occurrence filtering across all samples")
    # format prop filtered output to create dataframe
    step_3_occ_clean_tibble <- 
      step_3_occ_clean %>%
        as_tibble() %>%
        dplyr::select(-Taxon) %>%
        filter(., reads > 0) %>%
        filter(., !sample %in% blank_samples)
    
    ## Add to ASV_nested
    ASV_nested <- ASV_nested %>%
      mutate(., Step3_tibble = list(step_3_occ_clean_tibble))
    
    ## Summary_file_3
    ASV_summary <- ASV_nested %>%
      transmute(Miseq_region, Summary_3 = purrr::map(Step3_tibble, ~ how.many(ASVtable = ., round = "3_NC_occ.filt"))) %>%
      left_join(ASV_summary) %>%
      mutate(Summary = map2(Summary, Summary_3, bind_rows)) %>%
      dplyr::select(-Summary_3)
    
    ASV_summary$Summary
  } else {
    message("Either no negative controls, not running occurrence filtering")
    }
```

### Step 3.5 No negative control filtering

Save step2 dataframe as step3_tibble, for next steps of filtering.

```{r no-nc-filt, echo=FALSE, warning=FALSE, message=FALSE}
if (is.null(ASV_nested$'negative control'[[1]]) |
  exists("no_associated_samples") |
  config_file$step3_nc_filt=="none" |
  !exists("negative_long") |
  (exists("negative_long") && nrow(negative_long) == 0)) {
        message("No negative controls, copying positive control output to step3")
# save previous steps for step3:
step_2 <- ASV_nested$Step2_cleaned_tibble[[1]]

## Add to ASV_nested
ASV_nested <- ASV_nested %>%
  mutate(., Step3_tibble = list(step_2))

## Summary_file_3
ASV_summary <- ASV_nested %>%
  transmute(Miseq_region, Summary_3 = purrr::map(Step3_tibble, ~ how.many(ASVtable = ., round = "3_no.filt"))) %>%
  left_join(ASV_summary) %>%
  mutate(Summary = map2(Summary, Summary_3, bind_rows)) %>%
  dplyr::select(-Summary_3)

ASV_summary$Summary
  } else {
    message("Negative controls already used for filtering")
    }

```

### Step 3.6 Create Sample associated negative control flag:

```{r neg-control-flags, echo=FALSE, warning=FALSE, message=FALSE}
if (!is.null(ASV_nested$'negative control'[[1]]) && !exists("no_associated_samples") && nrow(negative_long) > 0) {
  message("Creating table to merge for sample nc flag")
  if (!is.null(step3_associated_nc)) {
  # keep negative flag id's only:
  neg_flags_to_add <- merged_table_long_v2 %>% 
    filter(!is.na(neg_flag)) %>% 
    select("neg_flag", "sample")
  
  # Add flags to samples for negative controls associated:
  neg_flags_to_add_1 <- blank_lookup %>% 
    separate_rows(sample, sep = "[,]+") %>% 
    left_join(neg_flags_to_add) %>%
    group_by(all_associated_samples) %>%
    summarize(sample = paste(sample, collapse = ","),
              neg_flag = paste(neg_flag, collapse = ","),
              .groups = "drop") %>%
    mutate(sample_nc_flag=ifelse(str_detect(neg_flag, "3"),"2","1")) %>%
    separate_rows(all_associated_samples, sep = "[,]+") %>%
    select(all_associated_samples,sample_nc_flag) %>%
    dplyr::rename(sample=all_associated_samples)
  
  } else {
    # filter out only negative controls from merged table for flags
    neg_flags_to_add <- merged_table_long_v2 %>% 
      filter(!is.na(neg_flag)) %>% 
      select("neg_flag", "sample") %>%
      summarise(sample_nc_flag=ifelse(str_detect(neg_flag, "3"),"2","1")) %>%
      summarise(sample_nc_flag=max(sample_nc_flag))
    
    # add sample id and merge
    neg_flags_to_add_1 <- merged_table_long_v2 %>%
      filter(samp_category=="sample") %>%
      mutate(sample_nc_flag=neg_flags_to_add$sample_nc_flag) %>%
      select(sample,sample_nc_flag)
  } 
 } else {
  message("No sample nc flag because no negative samples associated were sequenced")
 }

```  

Add sample_nc_flag to step3_tibble

```{r add-nc-flag-to-nested, echo=FALSE, warning=FALSE, message=FALSE}
if (!is.null(ASV_nested$'negative control'[[1]])&& !exists("no_associated_samples") && nrow(negative_long) > 0) {
  message("Creating table to merge for sample nc flag")
## Add to ASV_nested step3_tibble and samples df
ASV_nested <- ASV_nested %>%
  mutate(
    Step3_tibble = purrr::map( # add filtered out contaminants df to nested dataframe
      Step3_tibble,
      ~ left_join(., hash_key) %>%
        left_join(neg_flags_to_add_1, by="sample")),
    sample = purrr::map( # add filtered out contaminants df to nested dataframe
      sample,
      ~ left_join(., hash_key) %>%
        left_join(neg_flags_to_add_1, by="sample"))
  )
 } else {
  message("No sample nc flag because no negative samples")
 }
```  


## Step 4: Remove the positive and negative control samples.

This is already done in the nested file format (Step3_tibble does not have negative controls)

## Step 5: Cleaning Process 2: **Remove Likely Lab Contaminants**

* Step 5.1: Remove any likely lab contaminants
* Step 5.2: Filter reads by length (if 12S non-target seqs included)
* Step 5.3: Remove off-target domains

### Step 5.1 Remove known lab contaminants

```{r remove-lab-contam, echo=FALSE, warning=FALSE, message=FALSE}
## 1. Combine samples and reads with taxonomy to filter:
step_3_final_output<-ASV_nested$Step3_tibble[[1]] %>% # previous output
  left_join(.,hash_key) %>%
  # Create a logical vector of FALSEs
  mutate(known_contaminant=rep(FALSE, length(sample)))
     
## 2. Loop through the rules in your contaminant dataframe
for (i in seq_len(nrow(step5_species_to_remove))) {
       level <- step5_species_to_remove$taxa_level[i]
       id <- step5_species_to_remove$taxa_id[i]
       # Update the logical vector: if it was already TRUE, keep it TRUE;
       # If logical vector was FALSE, keep FALSE, if NA, keep FALSE by using %in%
       # If TRUE already and next value is NA (bc taxa_level not id'd), keep TRUE
       step_3_final_output <- step_3_final_output %>%
         mutate(known_contaminant=known_contaminant | 
                  (step_3_final_output[[level]] %in% id))
}
  
## 3. Output contaminant species:
known_contam_to_remove <- step_3_final_output %>%
  filter(known_contaminant==TRUE)
# view contaminants:
known_contam_to_remove
# save known comtaminants found:
write_csv(known_contam_to_remove, file = here(config_file$output_path,config_file$region,"Output_csv","step5_known_contam_removed.csv"))

## Add to ASV_nested
ASV_nested <- ASV_nested %>%
  mutate(
    Step5_remove_known_contaminants_tibble = purrr::map( # add filtered out contaminants df to nested dataframe
      Step3_tibble,
      ~ left_join(., hash_key) %>%
        filter(., !seq_id %in% known_contam_to_remove$seq_id)
    )
  )

## Summary_file_5
ASV_summary <- ASV_nested %>%
  transmute(Miseq_region,
    Summary_5 = purrr::map( 
      Step5_remove_known_contaminants_tibble, # add to summary file
      ~ how.many(ASVtable = ., round = "5_labcontaminants")
    )
  ) %>%
  left_join(ASV_summary) %>%
  mutate(Summary = map2(Summary, Summary_5, bind_rows)) %>%
  dplyr::select(-Summary_5)


ASV_summary$Summary

```

### Step 5.2 filter by sequence length

- Filter out by length if indicated in config file (seq_length_filt=="Yes")
  - seq_length_max indicates how to filter

Create dataframe of ASVs to remove:

```{r filter-by-seq-length, echo=FALSE, warning=FALSE}
if (config_file$seq_length_filt=="Yes") {
  message("Filtering ASVs by sequence length")
# function to output sequence length from each file:
  merged_seq_length_filt <- merged_table_long_v2 %>%
      filter(samp_category=="sample") %>% # filter out controls
      left_join(seq_len_df, by="seq_id") %>%
      mutate(off_target_length=case_when(
        config_file$seq_length_filt=="Yes" & 
          Seq_length > config_file$seq_length_max ~ "Off Target", 
        TRUE ~ "Not off target"
      ))

# save file of sequence length
write_csv(merged_seq_length_filt %>% filter(off_target_length=="Off Target"), 
            here(config_file$output_path,config_file$region,"Output_csv","Step5_sequence_length_filt.csv"))
# list of ASVs to remove based on sequence length
seqs_to_keep<-seq_len_df %>% 
  # filter out anything greater than max length
    filter(Seq_length<=config_file$seq_length_max)

# Add changes to ASV_nested
ASV_nested <- ASV_nested %>%
  mutate(
    # add filtered out contaminants df to nested dataframe
    Step5_seqlength_filt_tibble = purrr::map( 
      Step5_remove_known_contaminants_tibble,
      ~ left_join(., hash_key) %>%
        filter(., seq_id %in% seqs_to_keep$seq_id)
    )
  )

# add to Summary file
ASV_summary <- ASV_nested %>%
  transmute(Miseq_region,
    Summary_5_seqLength = purrr::map( 
      Step5_seqlength_filt_tibble, # add to summary file
      ~ how.many(ASVtable = ., round = "5_seqlength")
    )
  ) %>%
  left_join(ASV_summary) %>%
  mutate(Summary = map2(Summary, Summary_5_seqLength, bind_rows)) %>%
  dplyr::select(-Summary_5_seqLength)

ASV_summary$Summary

  } else { # If any other region besides 12S
    message("No sequence length filtering needed")
    
  }

```


### Step 5.3: Remove off-target domains

- 12S: 'Bacteria','Archaea'
- 18Sv4/18Sv9: 'Bacteria','Archaea'
- COI: 'Bacteria','Archaea'
- 16Sv4:
- ITS1: 'Bacteria','Archaea'

Create dataframe of these off-target taxa:
```{r load-off-target-taxa, warning=FALSE}
# Domains to filter out:
metabar_off_target<-data.frame(
                        metabar_region = c("12S","12S","18Sv4","18Sv4","18Sv9","18Sv9",
                                           "COI","COI","ITS1","ITS1"),
                        Domain=c("Bacteria","Archaea","Bacteria","Archaea","Bacteria","Archaea",
                                 "Bacteria","Archaea","Bacteria","Archaea")) %>%
  filter(metabar_region==config_file$region)
# Taxonomy to filter everything else out by:
metabar_taxa_only<-data.frame(metabar_region = "12S", taxa_level= "Phylum", taxa_id="Chordata") %>%
  filter(metabar_region==config_file$region)
```

Filter out off-target based on region:

```{r remove-off-target-taxa, echo=FALSE, warning=FALSE}
if (config_file$seq_length_filt=="Yes") {
  message("Starting from Step5_seqlength_filt_tibble")
      # remove off-target contaminants:
    ASV_nested <- ASV_nested %>%
      mutate(
        Step5_non_target = purrr::map( # add filtered out contaminants df to nested dataframe
          Step5_seqlength_filt_tibble,
          function(df) {
            # 1. Expand Taxonomy
            df_expanded <- df
            # left_join(., seq_id) %>%
            # 2. Filter out OFF-TARGET Domains
            # Handles multiple domain entries (e.g., c("Eukaryota", "Archaea"))
            if (!is.null(metabar_off_target$Domain)) {
              df_expanded <- df_expanded %>%
                filter(!Domain %in% metabar_off_target$Domain)
              }
            # 3. Filter by TARGET TAXA (Only if the reference dataframe is not empty)
            if (nrow(metabar_taxa_only) > 0) {
              target_levels <- unique(metabar_taxa_only$taxa_level)
              allowed_ids   <- metabar_taxa_only$taxa_id
          
            df_expanded <- df_expanded %>%
              filter(if_any(all_of(target_levels), ~ .x %in% allowed_ids))
            }
            return(df_expanded)
            }
          )
      )
 
  } else { # If any other region besides 12S
    message("Starting from Step5_remove_known_contaminants_tibble")
    # remove off-target contaminants:
    ASV_nested <- ASV_nested %>%
      mutate(
        Step5_non_target = purrr::map( # add filtered out contaminants df to nested dataframe
          Step5_remove_known_contaminants_tibble,
          function(df) {
            # 1. Expand Taxonomy
            df_expanded <- df
            # left_join(., seq_id) %>%
            # 2. Filter out OFF-TARGET Domains
            # Handles multiple domain entries (e.g., c("Eukaryota", "Archaea"))
            if (!is.null(metabar_off_target$Domain)) {
              df_expanded <- df_expanded %>%
                filter(!Domain %in% metabar_off_target$Domain)
              }
            # 3. Filter by TARGET TAXA (Only if the reference dataframe is not empty)
            if (nrow(metabar_taxa_only) > 0) {
              target_levels <- unique(metabar_taxa_only$taxa_level)
              allowed_ids   <- metabar_taxa_only$taxa_id
          
            df_expanded <- df_expanded %>%
              filter(if_any(all_of(target_levels), ~ .x %in% allowed_ids))
            }
            return(df_expanded)
            }
          )
      )
    }


```

Add to ASV_summary file:

```{r add-off-target-to-summary, echo=FALSE, warning=FALSE}
# add to Summary file
ASV_summary <- ASV_nested %>%
  transmute(Miseq_region,
    Summary_5_offTarget = purrr::map( 
      Step5_non_target, # add to summary file
      ~ how.many(ASVtable = ., round = "5_offTarget")
    )
  ) %>%
  left_join(ASV_summary) %>%
  mutate(Summary = map2(Summary, Summary_5_offTarget, bind_rows)) %>%
  dplyr::select(-Summary_5_offTarget)

print(tail(ASV_summary$Summary,10))

```

## Final formatting of files before moving to Part II

```{r final-formatting-output, echo=FALSE, warning=FALSE, message=FALSE}
# Save summary and nested data
saveRDS(ASV_nested, file = here(config_file$output_path,config_file$region,"Output_R","ASV_nested_final_partI.RDS"))
saveRDS(ASV_summary, file = here(config_file$output_path,config_file$region,"Output_R","ASV_summary_final_partI.RDS"))

# save final files
# ASVs and fasta combined
partI_ASV_final_output <- ASV_nested %>%
  dplyr::select(Step5_remove_known_contaminants_tibble) %>% 
    unnest(Step5_remove_known_contaminants_tibble) %>%
  unique() %>%
  dplyr::select(-c(Domain:Species)) %>%
    pivot_wider( # pivot wider so read counts are now below each sample name
      names_from = sample,
      values_from = reads,
      values_fill = list(reads = 0)
    )
# metadata file:
if (!is.null(ASV_nested$'negative control'[[1]]) && !exists("no_associated_samples")  && nrow(negative_long) > 0) {
  message("Merge metadata for sample nc flag")
  partI_meta_final_output <- merged_table_long_v2 %>%
    # remove ASV info:
    dplyr::select(-c(ASV:Sequence, reads)) %>%
    # remove duplicates from ASV info
    unique() %>% 
    # add sample flags for associated negative controls
    left_join(neg_flags_to_add_1, by="sample") %>%
    # add sample metadata:
    left_join(metadata %>% dplyr::select(-c(perc_reads_retained, neg_cont_type, samp_category)), 
              by=join_by("sample"=="samp_name"))
} else {
  partI_meta_final_output <- merged_table_long_v2 %>%
    # remove ASV info:
    dplyr::select(-c(ASV:Sequence, reads)) %>%
    # remove duplicates from ASV info
    unique()  %>%
    # add sample metadata:
    left_join(metadata %>% dplyr::select(-c(perc_reads_retained, neg_cont_type, samp_category)), 
              by=join_by("sample"=="samp_name"))
}

# add positive control flag if 18Sv4:
if(exists("OME_18Sv4_pos")) { 
  partI_meta_final_output <- partI_meta_final_output %>% 
    left_join(OME_18Sv4_pos %>% select(sample, Pos_pass_filters, pos_flag), by="sample")
} else {
  message("skip adding positive control flags")
}

# save metadata and asv file:
write_csv(partI_ASV_final_output, file = here(config_file$output_path,config_file$region,"Output_csv","asv_final_partI.csv"))
write_csv(partI_meta_final_output, file = here(config_file$output_path,config_file$region,"Output_csv","meta_final_partI.csv"))

```

#### Software and package versions used for file:

```{r warning=FALSE}
print(sessionInfo(), locale = FALSE)
```
