---
title: "Run1 OME Taxonomy Comparison Decontamination"
author: "Sam Setta"
date: "`r Sys.Date()`"
output: html_document
---

## Goal of Script

Set filtering method for ASV decontamination, steps we decided on:

**Part I Processing steps - THIS SCRIPT**

 * Step 1: Merge reads only (DONE)
 * Step 2: Tag jumping/index hopping
 * Step 3: Remove negative control contaminants.
 * Step 4: Remove the positive and negative control samples.
 * Step 5: Remove known contaminants (human, food products, pets, lab controls, etc.). Remove off-targets (i.e. taxonomic: bacteria/archaea in the MiFish; or length based). Leave in unknowns.

Produce Part I filtered output file

Following: https://github.com/zjgold/gruinard_decon

More details on how each works copied for the Readme:

**Estimation of Tag-jumping or Index Hopping**

With this procedure, we substract the proportion of reads that jumped into control samples from each environmental sample. First we determine which ASVs came from controls vs environmental samples. Next we create a vector of ASVs in positive controls. We then calculate what proportion of the reads found in the positive controls are found in environment samples. Next, we substract the composition of the positive controls from the environment samples. The idea behind this procedure is that we now know how many reads from each ASV appeared in the controls. These come from 2 processes: sequences we know should appear in the positive controls, and sequences that have hopped from the environment to the positive controls.

Load libraries

```{r setup, warning=FALSE, message=FALSE}
# install libraries
# BiocManager::install(c("tidyverse", "lubridate", "RColorBrewer", "ggsci", "dplyr", "phyloseq", "knitr", "readxl","ggplot2","NatParksPalettes","purrr", "ggvenn", "zetadiv"))

library("tidyverse")
# library("ggplot2") loads with tidyverse
# library("stringr") loads with tidyverse
# library("dplyr") loads with tidyverse
# library("tibble") loads with tidyverse
library("proxy")
library("reshape2")
library("microDecon")
library("knitr")
library("vegan")
library("phyloseq")
library("DESeq2")
library("ape")
library("RColorBrewer")
library("ggsci")
library("knitr")
library("readxl")
library("here")
# packages I don't think I'll use:
# library("ranacapa")
# library("plotly")
library("optparse")
# library("fitdistrplus")
# library("rstan")
# library("shinystan")
# library("bayesplot")
# library("broom")
library("configr")
library("Biostrings")
library("NatParksPalettes")
# library("zetadiv")
here()

source(here("Taxonomy_scripts","decontamination_utilities.R"))

```

## Set-Up for decontamination

### Load config file for processing.

```{r warning=FALSE, message=FALSE}
# load config file of variables for script
option_list <- list(
  make_option(c("-c", "--config"), type="character", default="decontam_pre-merge_12S.yaml",
              help="Path to configuration file [default %default]", metavar="FILE")
)

opt_parser <- OptionParser(option_list=option_list)
opt <- parse_args(opt_parser)
# check if yaml
is.yaml.file(opt$config)
eval.config.sections(opt$config)
config_file <- read.config(opt$config)
# what metabarcoding region are we working with?
config_file$region

```

### Load data

```{r warning=FALSE, message=FALSE}
# create output directories:
dir.create(here(config_file$output_path,"Output_csv"), recursive = TRUE)
dir.create(here(config_file$output_path,"Output_R"), recursive = TRUE)
dir.create(here(config_file$output_path,"Output_plots"), recursive = TRUE)
# load metadata file: (period in sample name rather than underscore)
metadata_file <- read_csv(
  config_file$input_meta_path
) 
  
# load and format asv table: (underscore in sample name rather than period)
revamp_asv_table_1 <- read.table(
  config_file$input_revamp_ASV_path_1,
  header = 1,
  sep = "\t",
  stringsAsFactors = F
)
# Remove MP_ from column head
colnames(revamp_asv_table_1) <-
  gsub("MP_", "", colnames(revamp_asv_table_1))
# load and format taxonomy file
revamp_taxonomy_table_1 <- read.table(
  config_file$input_revamp_taxonomy_path_1,
  header = 1,
  sep = "\t",
  stringsAsFactors = F
)
# load fasta file
revamp_fasta_table_1 <- readDNAStringSet(config_file$input_revamp_fasta_path_1,
  format =
    "fasta"
)
# format fasta file
fasta_df <- as.data.frame(revamp_fasta_table_1) %>%
  as_tibble(rownames = "ASV") %>%
  dplyr::rename("Sequence" = `x`)

```

### If 12S remove larger length region (bacteria) because we are targeting groups with shorter 12S

```{r warning=FALSE}
if (config_file$region=="12S") {
  message("Filtering ASVs by sequence length")
# function to output sequence length from each file:
  seq_len_df<-as.data.frame(fasta.seqlengths(config_file$input_revamp_fasta_path_1,
               nrec=-1L, skip=0L, 
               seek.first.rec=FALSE,
               seqtype="B", 
               use.names=TRUE)) %>%
    rownames_to_column(., var = "Feature.ID") %>%
    rename_with(~ "Seq_length",2) %>%
    mutate(seq_length_flag=case_when(Seq_length <= 210 ~ "Target length",
                                     TRUE ~ "Seq Length longer than 210bp"))
  # save file of sequence length
  write_csv(seq_len_df, 
            here(config_file$output_path,"Output_csv","Step0.12S.sequence_length_filt.csv"))
  # dataframe to count total reads, ASVs, samples before and afer filtering:
  filt_seq_length<-revamp_asv_table_1 %>% 
    pivot_longer(!x, names_to = "sample", values_to = "reads") %>%
    ungroup() %>% 
    left_join(seq_len_df, by=join_by("x"=="Feature.ID"))
  # ASVs filtered out:
  ASVs_gt_210bp<-filt_seq_length %>% filter(Seq_length<=210)
  # Add counts before filtering:
  read_summary_original<- filt_seq_length %>% 
    dplyr::summarise(nsamples = n_distinct(sample),
                     nHashes = n_distinct(x),
                     nReads = sum(reads), 
                     Stage = "Step_0.all.reads") %>% 
    gather(starts_with("n"), value = "number", key = "Stat")
  # Add counts after filtering:
  read_summary_filtered<-  filt_seq_length %>%
    filter(Seq_length<=210) %>%
    dplyr::summarise(nsamples = n_distinct(sample),
                     nHashes = n_distinct(x),
                     nReads = sum(reads), 
                     Stage = "Step_0.12S.seq.filtered") %>% 
    gather(starts_with("n"), value = "number", key = "Stat")
  
  # add two summary files:
  filt_read_summary<-rbind(read_summary_original,read_summary_filtered)
  # save file:
  write_csv(filt_read_summary, 
            here(config_file$output_path,"Output_csv","Step0.12S.summary_filt.csv"))
  
  # filter fasta table for later processing:
  fasta_df <- fasta_df %>% filter(ASV %in% ASVs_gt_210bp$x)
  # filter asv table for later processing:
  revamp_asv_table_1 <- revamp_asv_table_1 %>% filter(x %in% ASVs_gt_210bp$x)
  
  } else { # If there isn't a positive control output message
    message("No sequence length filtering needed")
  }
```

### Format data

Metadata:

```{r warning=FALSE, message=FALSE}
# format metadata file
metadata <- metadata_file %>%
  mutate(region=case_when( # create region column to filter by for decontam.
        assay_name=="ssu16sv4v5_parada" ~ "16Sv4",
        assay_name=="ssu18sv4_stoeck" ~ "18Sv4",
        assay_name=="ssu18sv9_amaralzettler" ~ "18Sv9",
        assay_name=="COI_1835-2198_lerayfolmer" ~ "COI",
        assay_name=="ssu12sv5v6_mifish_u_sales" ~ "12S",
        assay_name=="lsu16s_2434-2571_kelly" ~ "16S_Kelly",
        assay_name=="ssu18sv8_machida" ~ "18S_Machida",
        assay_name=="ITS1_sterling" ~ "ITS1"
  )) %>%
  dplyr::mutate( # run1 only has one sample with cruise code (WCOA21), remove that pattern.
    samp_name = dplyr::case_when(
      seq_run_id == "pmel_ome_run1" ~ gsub(".WCOA21", "", samp_name),
      TRUE ~ samp_name
    )) %>%
  separate_wider_delim( # Separate out sample id to get cruise id
    samp_name,
    delim = ".",
    names = paste0("part", 1:3), # separating columns
    too_few = "align_end",
    too_many = "drop",
    cols_remove = FALSE
  ) %>%
  mutate( # if run1, add WCOA21 as cruise code, if not run1 add cruise code from part3 column
    cruise = dplyr::case_when(
      seq_run_id == "pmel_ome_run1" ~ "WCOA21",
      seq_run_id != "pmel_ome_run1" ~ part3,
      TRUE ~ NA
    )) %>%
  dplyr::select(-c(part1:part3)) %>% # remove additional columns that are no longer needed
  mutate(samp_name=gsub("\\-.*", "", samp_name)) %>% # remove the cruise code after dash, since revamp output doesn't have this (ASV table)
  mutate(samp_name=gsub("\\.", "_", samp_name)) %>% # substitute any periods with underscores to match revamp output (ASV table etc.)
  mutate(perc_reads_retained=(output_read_count/input_read_count)*100) %>% # calculate percent reads retained (from raw reads to final ASV assignment)
  filter(region %in% config_file$region) %>% # only keep region of interest
  mutate(Control_Type=case_when( # add column for blanks, positive controls, or samples
            str_detect(samp_name, "_NC")  ~ "Blank",
            str_detect(samp_name, "POSITIVE")  ~ "Positive",
            TRUE ~ "Sample"
            ),
          Sample_Control=case_when( # add column to separate control from environmental samples
            Control_Type=="Sample" ~ "Sample",
                                   Control_Type=="Blank" ~ "Control",
                                   Control_Type=="Positive" ~ "Control"
            ),
         samp_name = case_when(
           str_detect(samp_name, "POSITIVE$") ~ "POSITIVE",
           TRUE ~ samp_name # Keeps original value if the condition is not met
         )) 
```

ASV table:

```{r warning=FALSE, message=FALSE}
# Generate Hash.key (essentially taxonomy table)
revamp_merged_df <- revamp_asv_table_1 %>%
  dplyr::rename("ASV" = `x`) %>%
  dplyr::select("ASV", any_of(metadata$samp_name)) %>%
  left_join(revamp_taxonomy_table_1) %>%
  left_join(fasta_df) %>%
  unite( # unite taxonomy into one column separated by semicolon
    .,
    "sum.taxonomy",
    c(
      "Kingdom",
      "Phylum",
      "Class",
      "Order",
      "Family",
      "Genus",
      "Species"
    ),
    sep = ";",
    remove = F
  )

# Keep only ASV and taxonomy columns
hash.key <- revamp_merged_df %>%
  dplyr::select(ASV, sum.taxonomy)

# save metabarcoding region id inmerged dataframe:
revamp_merged_df$Miseq_region <- config_file$region
```

### Merge tables (if using more than one and check format)

```{r warning=FALSE, message=FALSE}
# Merge All Tables (if using Anacapa taxonomy, but here, just based off revamp)
merged.table <- revamp_merged_df
# check output"
tail(merged.table)
head(merged.table)

```

### Convert to long format:

```{r warning=FALSE, message=FALSE}
### Format for Long Data
merged.table$ASV <- factor(merged.table$ASV)
merged.table$Miseq_region <- factor(merged.table$Miseq_region)

# save column names from merged.table:
columns <- colnames(merged.table)
# create list of column names to remove
remove <- c(
  "ASV",
  "sum.taxonomy",
  "Miseq_region",
  "Kingdom",
  "Phylum",
  "Class",
  "Order",
  "Family",
  "Genus",
  "Species",
  "Sequence"
)

# columns to keep via inverse filtering (all those not found in the 'remove' list)
gathercols <- columns[!columns %in% remove]

# create long dataframe by using only the sample IDs (those not found in 'remove' or only those found in 'gathercols')
merged.table_long <- merged.table %>%
  # select(-remove) %>%
  # column_to_rownames("ASV") %>%
  pivot_longer(
    .,
    cols = all_of(gathercols),
    names_to = "sample",
    values_to = "reads"
  )

# save reads as numeric
merged.table_long$reads <- as.numeric(merged.table_long$reads)

merged.table_long %>% filter(str_detect(sample,"POSITIVE"))

```

## Step 1: Merge reads only (DONE)

Need to change this if using anacapa results.

## Step 2: Tag jumping/index hopping 

### 2.1 Nest the dataset by origin of ASVs:

#### Label ASVs identified in positive control, negative control, or Samples

Identify Positives, Negatives, and Samples

```{r warning=FALSE, message=FALSE}
# save percent reads retained as vector:
reads_retained_con<-metadata %>%
  dplyr::select("samp_name","perc_reads_retained") 

### New column that labels each ASV as from Positive (control) or Sample
merged.table_long_v2 <-
merged.table_long %>%
  mutate(Control_Type=case_when( # add column for blanks, positive controls, or samples
            str_detect(sample, "_NC")  ~ "Blank",
            str_detect(sample, "POSITIVE")  ~ "Positive",
            TRUE ~ "Sample"
            ),
          Sample_Control=case_when( # add column to separate control from environmental samples
            Control_Type=="Sample" ~ "Sample",
                                   Control_Type=="Blank" ~ "Control",
                                   Control_Type=="Positive" ~ "Control"
            ),
         Pos = case_when( # Pos column is either Positive or Sample
          Control_Type=="Positive" ~ "Positive",
          TRUE ~ "Sample"
            )
         ) %>%
  left_join(reads_retained_con, by = join_by("sample"=="samp_name")) %>% # add percent reads retained and control types
  filter(reads != 0) # removing empty sequences here bc long format, so just adds more computational time here to include.

```

Create positive control filters:

- % of raw reads retained greater than 30
- Total reads > 30000

```{r warning=FALSE, message=FALSE}
# Check if there is a positive control in samples
if ("Positive" %in% merged.table_long_v2$Control_Type) {
  message("Positive control found — running code")
    # create a dataframe to use for positive control tag jumping decision later:
    pass_pos_filter_df <- merged.table_long_v2 %>%
      filter(Pos=="Positive") %>%
      group_by(sample, perc_reads_retained) %>%
      summarise(TotalReadsperSample = sum(reads)) %>%
      mutate(Pos_pass_filters=case_when(
        perc_reads_retained > 30 & TotalReadsperSample > 30000 ~ "pass",
        perc_reads_retained > 30 & TotalReadsperSample <= 30000 ~ "TooFewReads",
        perc_reads_retained <= 30 & TotalReadsperSample > 30000 ~ "LowReadsRetained",
        perc_reads_retained <= 30 & TotalReadsperSample <= 30000 ~ "fail")) %>%
      dplyr::select(c(sample,Pos_pass_filters))
    
    # merge df and pos_filter_df:
    merged.table_long_v2 <- merged.table_long_v2 %>%
      left_join(pass_pos_filter_df)

  } else { # If there isn't a positive control output message
    message("Positive control not found — skipping analysis")
  }
```

#### ASVs in positive controls

Beginning of code below will skip analysis if there are no positive controls present in the data.

```{r warning=FALSE, message=FALSE}
# Check if there is a positive control in samples
if ("Positive" %in% merged.table_long_v2$Control_Type) {
  message("Positive control found — running code")
  ## Summarise total number of reads per ASV
  all.seqs.in.positives <- merged.table_long_v2 %>%
    filter(Control_Type == "Positive") %>%
    dplyr::group_by(ASV) %>%
    dplyr::summarise(tot = sum(reads)) %>%
    arrange(desc(tot)) 
  
  ## Pull in associated taxonomy for positive ASVs
  pos.contam.species <- all.seqs.in.positives %>%
    left_join(hash.key, by="ASV") %>%
    as_tibble() 
  
  ## Save ASVs, taxonomy and read counts for positive control samples
  write_csv(pos.contam.species, file = here(config_file$output_path,"Output_csv","step2_pos.contam.species.csv")
            )
} else { # If there isn't a positive control output message
  message("Positive control not found — skipping analysis")
}
```

#### ASVs in negative controls.

Beginning of code below will skip analysis if there are no negative controls present in the data.

```{r warning=FALSE, message=FALSE}
# Check if there is a positive control in samples
if ("Blank" %in% merged.table_long_v2$Control_Type) {
  message("Negative control found — running code")

  ## Summarise total number of reads per ASV
  all.seqs.in.blanks <- merged.table_long_v2 %>%
    filter(Control_Type == "Blank") %>%
    dplyr::group_by(ASV) %>%
    dplyr::summarise(tot = sum(reads)) %>%
    arrange(desc(tot))
  
  ## Pull in associated taxonomy for positive ASVs
  blank.contam.species <- all.seqs.in.blanks %>%
    left_join(hash.key, by="ASV") %>%
    as_tibble() # OLDE CODE, already tibble if counts are included!
  
  ## Save ASVs, taxonomy and read counts for negative control samples
  write.csv(blank.contam.species, here(config_file$output_path,"Output_csv","step2_blank.contam.species.csv"))
} else { # If there isn't a positive control output message
  message("Negative control not found — skipping analysis")
}
```

#### Visualize Read Counts Across Samples for Barcode_1

```{r warning=FALSE, message=FALSE}
plot_1 <- merged.table_long_v2 %>%
  group_by(sample) %>%
  # filter(., Miseq_region == barcode_1) %>%
  mutate(TotalReadsperSample = sum(reads)) %>%
  arrange(desc(TotalReadsperSample)) %>%
  ggplot(., aes(x = sample, y = TotalReadsperSample, color = Control_Type)) +
  geom_point() +
  ggtitle("Read Count Across Samples") +
  theme(axis.text.x = element_text(angle = 90))

ggsave(
  plot = plot_1,
  here(config_file$output_path,"Output_plots","step1_Sample_Read_Depth_Barcode_1.png"),
  device = "png",
  width = 12,
  height = 8,
  units = "in"
)

```

#### Generate nested ASV table

```{r warning=FALSE, message=FALSE}
## Generated ASV.nested
ASV.nested<- merged.table_long_v2 %>%
    dplyr::group_by(Miseq_region, Pos) %>%
    nest() %>%
    pivot_wider(names_from = Pos, values_from = data)
  
### Summary.file.1 
ASV.summary <- ASV.nested %>%
    ungroup() %>%
    dplyr::mutate(., Miseq_region, Summary = purrr::map(Sample, ~ how.many(ASVtable = ., round = 0)), .keep = "none")
  ASV.summary$Summary

```

### 2.2: Model the composition of the positive controls of each run

```{r warning=FALSE, message=FALSE}
# Only run tag jumping in and out of positive controls if positive control in samples:
if (("Positive" %in% merged.table_long_v2$Control_Type) &&
     ("pass" %in% merged.table_long_v2$Pos_pass_filters) ) {
  message("Positive control found & passed filters — running code")
  #### Jumping vector out of Positive
ASV.nested <- ASV.nested %>%
    mutate(contamout.tibble = purrr::map(Sample, function(.x) { # started new nested dataframe in R
      df <- .x %>%
        ungroup() %>%
        group_by(sample) %>%
        mutate(TotalReadsperSample = sum(reads))  %>% # calculate total reads per sample
        mutate(proportion = reads / TotalReadsperSample) %>% # calculate proportion of reads per sample
    # remove known positive control species and other higher level taxonomic:
    filter(., sum.taxonomy == config_file$step1_positive_species_taxonomic_path | # with species
                 !!sym(config_file$step1_positive_lowest_taxonomic_lvl) == config_file$step1_positive_lowest_taxonomic_lvl_id) %>% # highest level taxonomy of positive 
        dplyr::select(-c(Kingdom:Species)) %>%
        ungroup() %>%
        group_by(ASV)
      # if positive contamination species are not found in samples (no rows, add NA values)
      if (nrow(df) == 0) {
      tibble(ASV = factor(NA, levels = levels(.x$ASV)), vector_contamination = NA_real_)
      # if positive contamination species are found in samples calculate maximum proportions for vector contamination
    } else {
      df %>%
        summarise(vector_contamination = max(proportion, na.rm = FALSE), .groups = "drop")
    }
    }))
 
  ### Jumping vector into Positive
   ASV.nested <- ASV.nested %>%
    mutate(contam.tibble = purrr::map(Positive, function(.x) { # create new nested dataframe using positive nested dataframe created above
      .x %>%
        ungroup() %>%
        group_by(sample) %>%
        mutate(TotalReadsperSample = sum(reads)) %>%  # calculate total reads per sample
        mutate(proportion = reads / TotalReadsperSample) %>% # calculate proportion of reads per sample
        group_by(ASV) %>% # this can be changed to sum.taxonomy
        dplyr::summarise(vector_contamination = max(proportion)) # calculate maximum proportions for vector contamination
    }))

  ### Vector Contamination in Barcode_1 visualization
  vc_plot_1<- rbind(ASV.nested$contam.tibble[[1]], ASV.nested$contamout.tibble[[1]]) %>% # combine vector contamination for sample (contamout.tibble) and positives (contam.tibble)
    as.data.frame() %>%
    ggplot(aes(x = vector_contamination)) + 
    geom_histogram() # y is the count of ASVs
  ggsave(
    plot = vc_plot_1,
    here(config_file$output_path,"Output_plots","step1_Vector_Contamination_Barcode_1.png"),
    device = "png",
    width = 12,
    height = 8,
    units = "in"
  )

} else { # If there isn't a positive control output message
  message("Positive control not found or didn't pass filters — skipping analysis")
}

ASV.nested$Positive


```

### 2.3: Substract the composition of the positive controls from the environment samples

- This is done ASV by ASV
- This could be translated to take the highest rate of index hopping observed for any non-positive control ASV and applied across all ASVs

```{r warning=FALSE, message=FALSE}
# Only run tag jumping in and out of positive controls if positive control in samples:
if (("Positive" %in% merged.table_long_v2$Control_Type) &&
     ("pass" %in% merged.table_long_v2$Pos_pass_filters)) {
  message("Positive control found & passed filters — running code")
  
  ## Find out what is the top contaminant across samples (not including known positive ASVs)
  max_vector_contamination <- rbind(ASV.nested$contam.tibble[[1]], 
                                    # contaminant proportions in positive controls
                                    ASV.nested$contamout.tibble[[1]]) %>% # contaminant proportions in negative controls
    left_join(hash.key) %>% # add taxonomy from taxonomy table
    separate_wider_delim(sum.taxonomy, delim = ";",
                         names = c("Kingdom","Phylum","Class","Order","Family","Genus","Species"), # Specify new column names
                         too_few = "align_start", # How to handle rows with fewer values
                         too_many = "drop", # How to handle rows with more values
      cols_remove = FALSE) %>%
    # remove known positive control species and genus:
    filter(., sum.taxonomy != config_file$step1_positive_species_taxonomic_path & # with genus
                 !!sym(config_file$step1_positive_lowest_taxonomic_lvl) != config_file$step1_positive_lowest_taxonomic_lvl_id 
           & Kingdom!="Unknown") %>% # with higher level taxonomy
    # include target taxonomy here
    arrange(desc(vector_contamination)) %>%
    dplyr::select(-c(Kingdom:Species)) %>%
    dplyr::slice(1) 

  ## Add to ASV.nested
  
  ASV.nested <- ASV.nested %>%
      ungroup() %>%
      mutate(Step2.cleaned.tibble = purrr::map(Sample, function(.x) {
        .x %>%
          dplyr::group_by(sample) %>%
          mutate(TotalReadsperSample = sum(reads)) %>% # calculate total reads per sample
          mutate(Updated_nReads = reads - (
            ceiling(
              max_vector_contamination$vector_contamination[[1]] * TotalReadsperSample
            )
          )) %>%
          filter(Updated_nReads > 0) %>%
          ungroup() %>%
          dplyr::select(sample, ASV, reads=Updated_nReads) # reads, Updated_nReads, TotalReadsperSample)
      }))
    
    ASVs_retained_step1 <- ASV.nested$Step2.cleaned.tibble[[1]]$ASV %>% unique()
  # which species were removed?: 
 ASV.nested$Sample[[1]] %>%
   filter(., !ASV %in% ASVs_retained_step1) %>%
   group_by(ASV, sum.taxonomy) %>%
   dplyr::summarise(n(), sum_reads =sum(reads) )%>% View()

  #### create summary file (Summary.file.2) 
  ASV.summary <- ASV.nested %>%
    dplyr::mutate(Miseq_region, Summary.2 = purrr::map(Step2.cleaned.tibble, ~ how.many(ASVtable = ., round = "1.Jump")), .keep = "none") %>%
    left_join(ASV.summary) %>% # use left join when there are many miseq runs to join
    mutate(Summary = map2(Summary, Summary.2, bind_rows)) %>%
    dplyr::select(-Summary.2)
  
  ASV.summary$Summary
  
  } else {
  message("Positive control not found or didn't pass filters - skipping analysis")
     ASV.nested <- ASV.nested %>%
      ungroup() %>%
      mutate(Step2.cleaned.tibble = purrr::map(Sample, function(.x) {
      .x %>%
        dplyr::select(sample, ASV, reads)
      }))
     
  ## Summary file so we have this step in our dataframe:
  ASV.summary <- ASV.nested %>%
    dplyr::mutate(Miseq_region, Summary.2 = purrr::map(Step2.cleaned.tibble, ~ how.many(ASVtable = ., round = "1.Jump")), .keep = "none") %>%
    left_join(ASV.summary) %>% # use left join when there are many miseq runs to join
    mutate(Summary = map2(Summary, Summary.2, bind_rows)) %>%
    dplyr::select(-Summary.2)
  
  ASV.summary$Summary
  
  }

```

## Step 3: Remove negative control contaminants

### Step 3.1: Investigate contaminants found in negative controls.

```{r warning=FALSE, message=FALSE}
# Only run negatice control analysis if there's a negative control in samples:
if ("Blank" %in% merged.table_long_v2$Control_Type) {
  message("Negative control found — running code")

  ## load in negative controls as separate df:
  neg_controls <- merged.table_long_v2 %>% # create dataframe of only samples that are negative controls.
    filter(Control_Type=="Blank") # only blanks (negative controls)

  ## Generate Long Data Frame of Negative Controls
  negative_long <- ASV.nested %>%
    dplyr::select(Step2.cleaned.tibble) %>% # pos contaminants removed tibble
    unnest(Step2.cleaned.tibble) %>% # unnest tibble to work with
    filter(sample %in% neg_controls$sample) # keep only negative control samples

  negative_long <- ASV.nested$Blanks[[1]] # save negative control long format data frame in nested table

  ### Reads per Contaminant:
  negative_reads <- negative_long %>%
      group_by(ASV) %>%
      dplyr::summarise(sum_reads_neg = sum(reads)) %>% # summarise total reads per ASV in blanks
      left_join(hash.key) %>% # join with taxonomy (to see taxa of contaminants)
      arrange(desc(sum_reads_neg)) # order ASVs with highest counts first
  
  negative_reads

} else {
  message("Negative control not found - skipping analysis")
  }
```

### Step 3.2: Calculations with contaminants in negative controls.

```{r warning=FALSE, message=FALSE}
# Only run negatice control analysis if there's a negative control in samples:
if ("Blank" %in% merged.table_long_v2$Control_Type) {
  message("Negative control found — running code")
  
  ### Generate Read Counts across all samples and negative controls:
  read_counts_step3 <- ASV.nested %>%
      dplyr::select(Step2.cleaned.tibble) %>% # pos contaminants removed tibble
      unnest(Step2.cleaned.tibble) %>% # This stage is a long table
      ungroup() %>% # unnest tibble to work with
      group_by(ASV) %>% 
      dplyr::summarise( # summarize read counts per ASV
        sum_reads = sum(reads), # Total reads per ASV
        mean_reads = mean(reads), # Mean reads per ASV
        median_reads = median(reads), # Median reads per ASV
        max_reads = max(reads), # Max reads per ASV
        sd_reads = sd(reads), # SD reads per ASV
        min_reads = min(reads) # Minimum reads per ASV
        ) %>%
    arrange(desc(sum_reads))
  
  ### Generate Occurrence Counts
  number_of_occurrences <-ASV.nested %>%
      dplyr::select(Step2.cleaned.tibble) %>%
      unnest(Step2.cleaned.tibble) %>% # This stage is a long table
      ungroup() %>%
      group_by(ASV) %>%
      dplyr::count() # Count number of times ASVs are found across all samples

  ### look at ASVs ASVs only found in negative controls (and not samples):
  setdiff(negative_long$ASV, number_of_occurrences$ASV)
  
  ### look at taxa shared between ASVs found in negative control and samples:
  intersect(negative_long$ASV, number_of_occurrences$ASV) %>% # which ASVs are shared?
    as_tibble() %>%
    dplyr::rename(ASV = value) %>%
    left_join(hash.key) # add taxa to ASVs 

  ## Create column of whether or not an ASV was found in blanks or samples
  number_of_occurrences <- number_of_occurrences %>%
    mutate(Contam = if_else(ASV %in% negative_long$ASV, "In Negatives", "In Samples Only")) %>%
    arrange(desc(n))

} else {
  message("Negative control not found - skipping analysis")
  }

```

### Step 3.3: Which Species are in the blanks and what are their occurrences & read counts

```{r warning=FALSE, message=FALSE}
# Only run negatice control analysis if there's a negative control in samples:
if ("Blank" %in% merged.table_long_v2$Control_Type) {
  message("Negative control found — running code")
  
  ### Which Species are in the blanks and what are their occurrences & read counts
   neg_comparison <- number_of_occurrences %>%
    filter(., Contam == "In Negatives") %>% # only look at ASV occurrences found in negative controls
    left_join(hash.key) %>% # merge ASV occurrence count with taxonomy
    left_join(negative_reads) %>% # join occurence with sum of negative reads per ASV
    left_join(read_counts_step3) %>% # join with summary of each ASV across all samples
    mutate(., max_n = max(number_of_occurrences$n))

  ### How would removing these ASVs change the number of results?
  ASV.nested %>%
    dplyr::select(Step2.cleaned.tibble) %>%
    unnest(Step2.cleaned.tibble) %>% # pull out previous filter step counts
    filter(., ASV %in% neg_comparison$ASV) %>% # select only ASVs in negative controls
    group_by(ASV) %>%
    dplyr::summarise(
      tot_reads_with_negASVs = sum(reads), # summarize reads for ASVs in neg control
      mean_reads_with_negASVs = mean(reads), # mean of reads for ASVs in neg control
      occurrences_with_negASVs = n_distinct(sample) # count how many times ASVs are found across all samples
    )

  ASV.nested %>%
    dplyr::select(Step2.cleaned.tibble) %>%
    unnest(Step2.cleaned.tibble) %>% # pull out previous filter step counts
    filter(., !ASV %in% neg_comparison$ASV) %>% # select only ASVs not in negative controls
    group_by(ASV) %>%
    dplyr::summarise(
      tot_reads_without_negASVs = sum(reads), # summarize reads for ASVs not found in blanks
      mean_reads_without_negASVs = mean(reads),  # mean of reads for ASVs not found in blanks
      occurrences_without_negASVs = n_distinct(sample) # count how many times ASVs are found across all samples
    )

} else {
  message("Negative control not found - skipping analysis")
}

```

Plot of negative controls counts

```{r warning=FALSE, message=FALSE}
# Only run negatice control analysis if there's a negative control in samples:
if ("Blank" %in% merged.table_long_v2$Control_Type) {
  message("Negative control found — running code")
  
  ### Plot Occurrences
  plot_occ <- number_of_occurrences %>% # plotting ASV occurence across samples
    filter(., n > 5) %>% # only plot ASVs that occur in at least 6 samples
    ggplot(., aes(x = reorder(ASV, n), y = n, fill = Contam)) + # fill with Contam status (in neg control or only in sample)
    geom_bar(
      stat =
        "identity"
    ) +
    facet_wrap(~Contam)

  ggsave(
    plot = plot_occ,
    here(config_file$output_path,"Output_plots",
         "step3_ASVs_from_Negative_Control_Occurrence_Plot.png"),
    device = "png",
    width = 12,
    height = 8,
    units = "in"
  )

} else {
  message("Negative control not found - skipping analysis")
}

```

### Step 3.3: Map List of ASVs Slated to be Removed

```{r warning=FALSE, message=FALSE}
# Only run negatice control analysis if there's a negative control in samples:
if ("Blank" %in% merged.table_long_v2$Control_Type) {
  message("Negative control found — running code")
  
  neg_comparison <- neg_comparison %>%
  mutate(
    .,
    Neg_remove = case_when(
      n < 10 ~ "To Remove", # if ASVs in negative controls only occur in less than 10 samples
      n / max_n > .9 ~ "To Keep", # Keep any ASVs found within top 10 % of max occurrence of all blank ASVs
      mean_reads - 2 * 1.96 * sd_reads > sum_reads_neg ~ "To Keep", # subtract the lower bounds of the 95% confidence interval of the normal distribution from the mean reads. If the mean reads are greater than this (within the normal distribution) and greater than the sum_reads_neg (reads in negative control) keep the ASV.
      TRUE ~ "To Remove"
    )
  )

} else {
  message("Negative control not found - skipping analysis")
}

```

### Step 3.4: Format and run for microDecon

Create fake negative control for microDecon if not already present, and define data for adding to nested ASV data frame.

```{r warning=FALSE, message=FALSE}
##### Identify blank samples
if (!is.null(ASV.nested$Blanks[[1]])) { # only run if negative controls in sample
  message("Negative control found — running code")
  blank_ASVs_1 <- ASV.nested$Blanks[[1]] %>% # Sample names and ASVs for blanks
    dplyr::select(sample, ASV, reads) # only keep selected columns
  #### Number of Blanks
  blank_num <- ASV.nested$Blanks[[1]]$sample %>%
    unique() %>%
    length()
  #### Columns to Ignore for Grouping
  blank_samples <- ASV.nested$Blanks[[1]]$sample %>% unique()
} else { # If there are no blanks in samples, skip this step.
  message("Negative control not found - skipping analysis")
  #### Columns to Ignore for Grouping
  blank_samples <- ASV.nested$Blanks[[1]]$sample %>% unique()
}
```

Format dataframe for microDecon

```{r warning=FALSE, message=FALSE}
# Only run if negative control found:
if (!is.null(ASV.nested$Blanks[[1]])) {
  message("Negative control found — running code")
  ##### Save previous steps to use for microDecon
  step.2_barcode1 <- ASV.nested$Step2.cleaned.tibble[[1]]
  #### Pivot to Wide format
  step.2_barcode1_wide <- step.2_barcode1 %>%
    dplyr::select(ASV, reads, sample) %>% # select only necessary columns
    pivot_wider( # pivot wider so read counts are now below each sample name
      names_from = sample,
      values_from = reads,
      values_fill = list(reads = 0)
    )

  #### Arrange Blank Samples to be the first 3 columns
  step.2_barcode1_wide <- step.2_barcode1_wide %>%
    left_join(hash.key) %>% # add taxonomy
    relocate(sum.taxonomy, .after = last_col()) %>% # move taxonomy to the end of the   dataframe
    relocate(blank_samples, .after = ASV) %>% # move blanks samples to the first three columns after the ASV column
    as.data.frame()

} else {
  message("Negative control not found - skipping analysis")
   ##### Save previous steps to use for microDecon
  step.2_barcode1 <- ASV.nested$Step2.cleaned.tibble[[1]]
}
```

Run (microDecontam)[https://github.com/donaldtmcknight/microDecon]

```{r warning=FALSE, message=FALSE}
# Only run if negative control found:
if (!is.null(ASV.nested$Blanks[[1]])) {
  message("Negative control found — running code")
  #### Number of Samples (for decontam code)
  sample_num<-ASV.nested$Step2.cleaned.tibble[[1]]$sample %>%
    unique() %>%
    length()

  # columns to remove 
  col_delete <- append(blank_samples, c("sum.taxonomy", "ASV", "Miseq_region"))

  #### Assign Groups -> should be replaced with pre-defined groups in the metadata file
  if (config_file$step5_som_level == "double") { # assign groups of samples (if nested=double)
  sample_counter <- colnames(step.2_barcode1_wide) %>%
    as.data.frame() %>%
    filter(., !(. %in% col_delete)) %>%
    mutate(., New_name = .) %>%
    left_join(metadata, join_by(New_name==samp_name)) %>%
    dplyr::select(New_name, sample_derived_from, rep) %>%
    distinct() %>%
    group_by(Site, Bio_rep) %>%
    dplyr::count()
} else {
    sample_counter <- colnames(step.2_barcode1_wide) %>%
      as.data.frame() %>%
      filter(., !(. %in% col_delete)) %>%
      mutate(., samp_name = .) %>%
      left_join(metadata, by = "samp_name") %>%
      group_by(Sample_Control) %>%
      dplyr::count()
  }

  step.3_barcode1_decon <- decon( # run microdecontam with settings from yaml:
    data = step.2_barcode1_wide,
    numb.blanks = blank_num,
    numb.ind = sample_counter$n,
    taxa = T,
    runs = config_file$step3_runs,
    thresh = config_file$step3_thresh,
    prop.thresh = config_file$step3_prop.thresh,
    regression = config_file$step3_regression
  )

  saveRDS(step.3_barcode1_decon,   here(config_file$output_path,"Output_R","step.3_barcode1_decon.RDS"))
        
  ### Convert to Long Data
  columns <- colnames(step.3_barcode1_decon$decon.table)
  remove <- c("ASV", "sum.taxonomy")
  # columns to gather in pivot longer
  gathercols <- columns[!columns %in% remove]

  # Save deconamtinated dataframe
  step.3_barcode1_decon_clean <-
    step.3_barcode1_decon$decon.table %>%
    pivot_longer(.,
      cols = gathercols,
      names_to = "sample",
      values_to = "reads"
    )
} else {
  step.3_barcode1_decon_clean <- step.2_barcode1 %>%
    left_join(hash.key, by="ASV")
  message("Negative control not found - skipping analysis")
}
  
```

### Step 3.5 Final Clean Up

```{r warning=FALSE, message=FALSE}
# format decontam output to create dataframe
step.3_barcode1_decon_clean_tibble <- 
  step.3_barcode1_decon_clean %>%
    as_tibble() %>%
    dplyr::select(-sum.taxonomy) %>%
    filter(., !sample %in% c("Mean.blank", "fake_asv_1")) %>% # remove the fake blanks if they were added above
    filter(., !ASV == "fake_asv_1") %>% # remove the fake ASV if they were added above
    filter(., reads > 0) %>%
    filter(., !sample %in% blank_samples)

##### Did any ASVs remain after microdecon?
if (!is.null(ASV.nested$Blanks[[1]])) { # Only run if negative control exists
  step.3_barcode1_decon_clean_tibble %>%
    filter(., ASV %in% negative_long$ASV)

  step.3_barcode1_decon_clean_tibble %>%
    left_join(hash.key) %>%
    filter(., sum.taxonomy %in% negative_long$sum.taxonomy) %>%
    filter(., !sample %in% blank_samples) %>%
    group_by(sum.taxonomy) %>%
    dplyr::summarise(sum_reads = sum(reads))
} else {
  message("Negative control not found - not running check on microdecon output")
}

## Add to ASV.nested
ASV.nested <- ASV.nested %>%
  mutate(., Step3.tibble = list(step.3_barcode1_decon_clean_tibble))

## Summary.file.3
ASV.summary <- ASV.nested %>%
  transmute(Miseq_region, Summary.3 = purrr::map(Step3.tibble, ~ how.many(ASVtable = ., round = "3.microDecon"))) %>%
  left_join(ASV.summary) %>%
  mutate(Summary = map2(Summary, Summary.3, bind_rows)) %>%
  dplyr::select(-Summary.3)

ASV.summary$Summary

```

## Step 4: Remove the positive and negative control samples.

```{r warning=FALSE, message=FALSE}
# Identify control samples
samples_only <- merged.table_long_v2 %>% # orginal merged sample dataframe
      filter(Control_Type!="Blank" & Control_Type!="Positive") # only samples (no negative or positive controls)

## Filter out negative and positive control samples
step.4_barcode1_removeCon_clean_tibble <- ASV.nested %>%
    dplyr::select(Step3.tibble) %>% # pos contaminants removed tibble
    unnest(Step3.tibble) %>% # unnest tibble to work with
    filter(sample %in% samples_only$sample) # keep only negative control samples

## Add to ASV.nested
ASV.nested <- ASV.nested %>%
  mutate(., Step4.tibble = list(step.4_barcode1_removeCon_clean_tibble))

## Summary.file.4
ASV.summary <- ASV.nested %>%
  transmute(Miseq_region, Summary.4 = purrr::map(Step4.tibble, ~ how.many(ASVtable = ., round = "4.removecontrols"))) %>%
  left_join(ASV.summary) %>%
  mutate(Summary = map2(Summary, Summary.4, bind_rows)) %>%
  dplyr::select(-Summary.4)

ASV.summary$Summary

```

## Step 5: Cleaning Process 2: **Remove Likely Lab Contaminants**

Load known contaminant taxonomy 

```{r warning=FALSE, message=FALSE}
### Filter by Taxonomic Paths
# This list can be expanded upon additional work to identify known contaminants within OME PMEL samples
if (is.null(config_file$step5_species_to_remove_path)) {
  step5_species_to_remove <- c(
    "Eukaryota;Chordata;Aves;Accipitriformes;Accipitridae;Harpagornis;Harpagornis moorei",
    "Eukaryota;Chordata;Mammalia;Primates;Hominidae;Pan;Pan paniscus",
    "Eukaryota;Chordata;Mammalia;Primates;NA;NA;NA",
    "Eukaryota;Chordata;Mammalia;Artiodactyla;Suidae;Sus;Sus scrofa",
    "Eukaryota;Chordata;Mammalia;Artiodactyla;Bovidae;Bos;Bos taurus",
    "Eukaryota;Chordata;Mammalia;Primates;Hominidae;Homo;Homo sapiens",
    "Eukaryota;Chordata;Aves;Galliformes;Phasianidae;Gallus;Gallus gallus",
    "Eukaryota;Chordata;Mammalia;Carnivora;Canidae;Canis;Canis lupus"
  )
} else {
  step5_species_to_remove <- read.csv(here(config_file$step5_species_to_remove_path))$sum.taxonomy
}

```

Remove contaminants from above

```{r warning=FALSE, message=FALSE}
### Code to visualize
known_contam_to_remove<-ASV.nested$Step4.tibble[[1]] %>% # previous output
  left_join(.,hash.key) %>%
  filter(., sum.taxonomy %in% step5_species_to_remove) # filter out contamination species
# view dataframe:
View(known_contam_to_remove)
# save known comtaminants found:
write_csv(known_contam_to_remove, file = here(config_file$output_path,"Output_csv","step5_known_contam_removed.csv"))

## Add to ASV.nested
ASV.nested <- ASV.nested %>%
  mutate(
    Step5.remove.known.contaminants.tibble = purrr::map( # add filtered out contaminants df to nested dataframe
      Step4.tibble,
      ~ left_join(., hash.key) %>%
        filter(., !sum.taxonomy %in% step5_species_to_remove)
    )
  )

## Summary.file.5
ASV.summary <- ASV.nested %>%
  transmute(Miseq_region,
    Summary.5 = purrr::map( 
      Step5.remove.known.contaminants.tibble, # add to summary file
      ~ how.many(ASVtable = ., round = "5.labcontaminants")
    )
  ) %>%
  left_join(ASV.summary) %>%
  mutate(Summary = map2(Summary, Summary.5, bind_rows)) %>%
  dplyr::select(-Summary.5)


ASV.summary$Summary

```

## Final formatting of files before moving to Part II

```{r warning=FALSE, message=FALSE}
# Save summary and nested data
saveRDS(ASV.nested, file = here(config_file$output_path,"Output_R","ASV.nested_final_partI.RDS"))
saveRDS(ASV.summary, file = here(config_file$output_path,"Output_R","ASV.summary_final_partI.RDS"))

# save final files
# ASVs and fasta combined
partI_ASV_final_output <- ASV.nested %>%
  dplyr::select(Step5.remove.known.contaminants.tibble) %>% 
    unnest(Step5.remove.known.contaminants.tibble) %>%
    pivot_wider( # pivot wider so read counts are now below each sample name
      names_from = sample,
      values_from = reads,
      values_fill = list(reads = 0)
    )
# metadata file:
partI_meta_final_output <- metadata %>%
  filter(samp_name %in% names(partI_ASV_final_output))

# save metadata and asv file:
write_csv(partI_ASV_final_output, file = here(config_file$output_path,"Output_csv","asv_final_partI.csv"))
write_csv(partI_meta_final_output, file = here(config_file$output_path,"Output_csv","meta_final_partI.csv"))

```

#### Software and package versions used for file:

```{r warning=FALSE}
print(sessionInfo(), locale = FALSE)
```
